---
title: "Advanced R for Econometricians"
subtitle: "Introduction"
author: 
  - "Martin Arnold"
  - "Martin Schmelzer"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
class: title-slide title-rintro center middle
```{r moon_reader, include=FALSE, eval = FALSE}
# setup for moon reader
# copy it to console
options(servr.interval = 0.5)
xaringan::inf_mr()
```

```{r xaringanExtra_progress-bar, echo = FALSE}
xaringanExtra::use_progress_bar(color = "#004c93", location = "bottom")
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css 
# https://github.com/gadenbuie/xaringanExtra
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(icons)

# pre-define icons
desktop <- icons::icon_style(icons::fontawesome$solid$desktop, fill = "#004c93")
checker_green <- icons::icon_style(icons::fontawesome$solid$check, fill = "green")
times_red <- icons::icon_style(icons::fontawesome$solid$times, fill = "red")
```


#  `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`
---
## Preliminaries

.pull-left-1[**Contact Information:**  ]

.pull-right-2[
<p> <span style="margin-bottom: -7px;">Martin Schmelzer </span><span><a href = "mailto: martin.schmelzer@outlook.com">martin.schmelzer@outlook.com</a></span></p> 

<p style="margin: 25px;"></p>

<p> <span style="margin-bottom: -7px;">Martin Arnold </span><span><a href = "mailto: Martin Arnold@vwl.uni-due.de">Martin.Arnold@vwl.uni-due.de</a>]</span></p> 
 


.pull-down[
Slides, exercises and announcements will be provided on: 
.center[
  https://moodle.uni-due.de/course/view.php?id=17174  
]
<br>
The password for the course is **"Hadley"**.  
<br>
<!--On Moodle you will find a link to [Datacamp](https://www.datacamp.com) which will give you free access to all their content. -->
<br>

]
???
Background? 
R and Rstudio installed? 
What do you know about R? 

---
## Grading

- You will have to work on a final group project which focuses on one or more of the topics discussed in the course. 

- The final grade is a weighted average of the grades for the group project (70%), its presentation (30%).

- Projects should be registered with us until June 16th. We will set up a form on Moodle.

---
### Outline

.pull-left[
1. Introduction
2. Data Visualisation with ggplot
3. Data Wrangling with dplyr
4. Databases
5. Advanced R Programming
  - Advanced Concepts
  - Functional Programming
  - Object Oriented Programming
]


.pull-right[
<ol start = "6">
  <li> Performance </li> 
  <ul>
    <li>Profiling and Benchmarking</li> 
    <li>Improving Performance</li>
  </ul>
  <li>Web Scraping</li>
  <li>Rmarkdown</li>
  <li>Git and Github</li>
</ol>
]




.pull-right[.pull-down-small_margin.font90[
#### Additional Topics for self-study

<ol start = "8">
  <li>data.table</li>
  <li>Shiny</li>
  <li>Writing Packages</li>
</ol>
]]
---

class: top, left
## About R
- **Advantages:**
  - R is a language and environment for statistical computing and graphics.
  - R is highly extensible. There is a massive set of packages for statistical modelling, machine learning, visualisation, as well as importing and manipulating data. Researchers in statistics and machine learning often publish an R package to accompany their articles.
  - R is free and open source. 
  - R provides tools for communicating your results. R packages make it easy to produce html or pdf reports, or create interactive websites.
  - R is designed to connect to high-performance programming languages like C, Fortran, and C++.

--

- **Disadavantages:**
  - It is not very fast.
  - It has an inconsistent syntax.

---
class: top, left
## Sources

- [Advanced R](https://adv-r.hadley.nz/) by Hadley Wickham
- [R for Data Science](https://r4ds.had.co.nz) by Hadley Wickham and Garrett Grolemund
- [The Art of R Programming](https://www.oreilly.com/library/view/the-art-of/9781593273842/) by Norman Matloff
- The [Use R!](https://www.springer.com/series/6991) Series by Springer
- The [R Series](https://www.crcpress.com/go/the-r-series) by CRC Press

---
class: top, left
## Packages

- In R the fundamental unit of shareable code is the package.
- Chances are that someone has already solved a problem that you’re working on, and you can benefit from their work by downloading their package.

#### The main sources for packages are:
  - CRAN (Comprehensive R Archive Network) with as of now ca. 15.000 available packages
  - GitHub (that is the place where packages are developed before getting published on CRAN) 
  - Bioconductor

---
class: top, left
## Installing Packages
.font80[
**How to install packages from CRAN?**

- Use `install.packages("MASS")` to install the package `MASS` from CRAN.
- Or using RStudio: Tools &rarr; Install Packages &rarr; `MASS`
]

--

.font80[
**How to install packages from Github?**

- Install the package `devtools` from CRAN.
- E.g., use `devtools::install_github("tidymodels/broom")` to install the `broom` package from Github.

- If you want to find out how a package works, search for

  - a vignette by the package authors 
  - a blog post about the package, e.g. on [R Bloggers](https://www.r-bloggers.com)
]

--

.font80[
.blockquote.exercise[

#### `r desktop` Example

```{r, echo = TRUE, eval = FALSE}
install.packages("broom")
library(broom)
vignette("broom")
```

]]

???

Which functionality does the broom package provide?
---
class: top, left

## Getting Help

If you have specific questions about your code or an error message look on
  - [StackOverflow](https://stackoverflow.com/)  
  - [RStudio Community](https://community.rstudio.com/) (if it is about RStudio or packages developed by RStudio)
  - the [GitHub](https://github.com/) repository of the package maintainer

The chances are high that somebody else has already asked the same question and it has already been solved. 

--

If you don't find an answer you can ask a question yourself. What sounds easy is harder then one would think. Most importantly you have to provide a [reproducible example](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example).

???
Bad questions: 
https://stackoverflow.com/questions/58302656/how-can-i-set-equals-function-in-objective-of-solver-in-r-python
https://stackoverflow.com/questions/58121493/update-a-coordinate-dataframe-in-shiny 

Good questions:  
https://stackoverflow.com/questions/38777337/ggplot-ribbon-cut-off-at-y-limits/38777929#38777929 
https://stackoverflow.com/questions/53241120/how-to-use-openblas-lapacke-together-with-rcpp

---
## RStudio

- RStudio is an integrated development environment (IDE) for R. 
- It provides some convenient functionality compared to the IDE that comes with R.

--

#### RStudio Projects 

- To work in a specific directory the base R way would be to use `setwd()`. This is problematic if you want to 
  share your code or work from different machines on the same project. 
- Use projects instead.

--

#### Workflow

- Instead of opening a new R session and use `setwd()` you open the project. The working path is 
automatically set to the folder where the `.Rproj` file lives in. 

---
#### Some Shortcuts
<p></p>

```{r echo = FALSE, eval = TRUE}
knitr::kable(data.frame(Shortcut = c("Alt + Shift + k", "Cmd/Ctrl + Shift + F10", "Cmd/Ctrl + Shift + S", "F1", "Ctrl + l"),
                        Action   = c("list all shortcuts", "restart R", "rerun current script", "help page", "clears the console"))
             , format = "html")
```

--

#### RStudio supports: 
- package development
- version control via git and svm 
- creation and compilation of dynamic documents 
- connections to external data bases 
- ... 

--

We will use some of this later on.  

---
class: top, left

## R Basics

- We start by recapitulating some basics you should be familiar with when working with R. 
- An in-depth treatment of this material can be found in the first chapter of
[Advanced R](https://adv-r.hadley.nz/vectors-chap.html#introduction) by Hadley Wickham. 


---
class: top, left

## R Basics: Data Types and Data Structures
.font90[
We assume that you know the basic data structures in R. 
  
```{r, echo = FALSE, eval = TRUE}
knitr::kable(
data.frame(c("1", "2", "n"), Homogeneous = c("Atomic Vector", "Matrix", "Array"), Heterogeneous = c("List", "Data Frame", "")),
col.names = c("Dimensions", "Homogeneous", "Heterogeneous"), format = "html")
```
<br>
] 

--

.font90[
You should also be familiar with the basic data types (character, double, integer, logical).

]

--
<br>

.font90[ .content-box-white[

**Task**

1. How is a list different from an atomic vector? 
2. How is a matrix different from a data frame?
3. How do you find out the type of an object?
4. How can you check if an object is of type character?

]]

???
**Answers Task**

1. list can store different data types
2. data frames can store different values with different types
3. class()
4. is.character()


```{r}
object.size(1:1000)
object.size(as.numeric(1:1000))

char_vector <- c("Hi", "there")
typeof(char_vector)
class(char_vector)

num_matrix <- matrix(rnorm(4), ncol = 2)
typeof(num_matrix)
class(num_matrix)
is.numeric(num_matrix)


plus <- function(a, b) {
  if(!is.numeric(a)) stop("a is not numeric")
  if(!is.numeric(b)) stop("b is not numeric")
  a + b
}

plus(1, "a")
```


---
class: top, left
## R Basics: Subsetting

You should know how to use the common subsetting operators  `[`, `[[` and `$`. 

Have you ever encountered `@`? 

--

<br>

<div class = "content-box-white">

  <strong> Task </strong>

  <ol>
    <li> What is the result of subsetting a vector with positive integers, negative integers, a logical vector, or a character vector? </li> 
    <li> What is the difference between <code> [ </code>,  <code> [[ </code>, and <code> $ </code> when applied to a list?  </li>
    <li>  When should you use <code> drop = FALSE </code> ? </li>
  </ol>

</div>


???

1. 

[] = always returns object of same class 
[[]] = can extract one element from list or data frame
list_a <- list(a = 2, 
     b = "hallo")
list_a["a"]
list_a[["a"]]
list_a$a

2. 
[ selects sub-lists. It always returns a list; if you use it with a single positive integer, it returns a list of length one. [[ selects an element within a list. $ is a convenient shorthand: x$y is equivalent to x[["y"]].

3. 
Use drop = FALSE if you are subsetting a matrix, array, or data frame and you want to preserve the original dimensions. You should almost always use it when subsetting inside a function.

```{r}

## Drop
m1 <- matrix(1:4, nrow = 2)
m2 <- matrix(1:4, nrow = 2)

matrix_prod <- function(A, B) {
  B %*% t(A)
}

matrix_prod(m1[1, ], m2)
matrix_prod(m1[1, , drop = FALSE], m2)

## Example for S4 class
N <- 200
x.vec <- as.vector(fGarch::garchSim(fGarch::garchSpec(rseed = 1985), n = N)[,1])
garch11 <- fGarch::garchFit(~ garch(1,1), data = x.vec, trace = FALSE)

isS4(garch11)

```


---
## R Basics: Control Flows

The basic control flows are conditionals (choices) and loops.

.content-box-white[
**Task**

1. What is the difference between `if` and `ifelse()`?
2. What will be the value of `y` in the following code if `x` is `TRUE`? What if `x` is `FALSE`? What if `x` is `NA`?
    ```{r, eval = FALSE}
y <- if(x) 3
    ```
3. Name three kind of loops that can be implemented in R! How do they differ?
]

???

```{r}
if("2") 3
if(2) 3

for(i in 1:3){
  print(i)
}

i <- 1
while(i < 4){
  print(i)
  i <- i + 1
}

i <- 1
repeat{
  print(i)
  i <- i + 1 
  if(i > 3) break
  
}
```

---
class: top, left

## R Basics: Functions

.font90.content-box-white[

**Task**

- Write a function that takes a number `x` and returns the square of `x`.  
]

--

#### Scoping

.font90[
- An important concept and also a source of trouble is scoping, the act of finding the value associated with a name. 
- R uses lexical scoping:  it looks up the values of names based on how a function is defined, not how it is called.
]

--

.blockquote.exercise[

#### `r desktop` Example: lexical scoping

.code90[
```{r}
f1 <- function() x
x  <- "global"
f2 <- function() {
  x <- "local_f2"
  f1()
}
f2()
```
]]
---
## Scoping Rules

R’s lexical scoping follows four primary rules:
- Name masking
- Functions versus variables
- A fresh start
- Dynamic lookup

???
Explaniation on the following slides


---
## Scoping Rules: Name Masking

.small[
.blockquote.exercise[ 

#### `r desktop` Example: What happens here?
```{r, eval = FALSE}
x <- 3
square_1 <- function() x^2
square_1()
```
]]

--

<br>

.small[
.blockquote.exercise[ 

#### `r desktop` Example: And here?

```{r}
x <- 3
square_2 <- function() { 
  x   <- 2
  foo <- function() x^2
  foo()
}
square_2()
```
]]
--
.small[
- Names defined inside a function mask names defined outside a function.
]

???
Example 1 works fine x will be not overwritten
Exmaple two -> x will be overwritten in the function

---
## Scoping Rules: Functions versus Variables

.small[
.blockquote.exercise[ 

#### `r desktop` Example: Let's make things more confusing. Can you predict the result?   

```{r}
x <- 3

square_2 <- function() { 
  x   <- 2
  foo <- function() x^2
  foo()
}

square_3 <- function(x) {
  square_2 <- 5
  square_2()
}

square_3()
```
]]

--

- When a function and a non-function share the same name, R ignores non-function objects in a function call.

---
## Scoping Rules: A Fresh Start

.blockquote.exercise[ 

#### `r desktop` Example: What is the result of the first call to `square_4()`? What will happen the second time?

```{r}
x <- 2
square_4 <- function() {
  (x <- x^2)
}

square_4()
square_4()
```
]

--

- Every time a function is called a new ***environment*** is created to host its execution. 
- A function has no way to tell what happened the last time it was run; each invocation is completely independent.

---
## Scoping Rules: Dynamic Lookup

.blockquote.exercise[ 

#### `r desktop` Example: Predict the result of the two calls to `f1()`. Compare it to the main rule of lexical scoping. 

```{r}
x  <- 1
f1 <- function() x
f1()

x <- 2 
f1()
```
]

--

- R looks for values when the function is run, not when the function is created. 

---
## Lazy Evaluation

.blockquote.exercise[ .font90[ .code90[

#### `r desktop` What could cause an error here?
```{r}
lazy_function <- function(x) {
  10
}

lazy_function()
```
]]]

--

.font90[

Why does this not induce an error?
]

--

.blockquote.exercise[ .font90[.code90[


#### `r desktop` Function arguments are lazily evaluated; they’re only evaluated if accessed.

```{r, eval = FALSE}
lazy_function <- function(x) {
  x
}

lazy_function()
```

]]]

---
## Style Guide

#### Object names

.font70[
- Variable and function names should be lowercase.
- Use an underscore (_) to separate words within a name (as an alternative you can use camel case, but be consistent).
- Variable names should be nouns.
- Function names should be verbs.
- Strive for names that are concise and meaningful (this is not easy!).
- Where possible, avoid using names of existing functions and variables. 
]

--

.pull-left[ .code80[ .blockquote.exercise[ 

#### `r checker_green` Good

```{r, eval = FALSE}
day_one
day_1
```
]]]

--

.pull-right[
.code80[
.blockquote.exercise[ 

#### `r times_red` Bad

```{r, eval = FALSE}
# Bad
first_day_of_the_month
dayone
djm1
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```
]]]

---
#### Spacing

.font80[
- Place spaces around all infix operators (`=`, `+`, `-`, `<-`, etc.) and `=` in a function call.
- Always put a space after a comma, and never before.

]

.pull-left[ .code80[ .blockquote.exercise[ 

#### `r checker_green` Good
```{r}

average <- mean(feet / 12 + inches, na.rm = TRUE)
```
]]]

.pull-right[ .code80[ .blockquote.exercise[ 

#### `r times_red` Bad

```{r}

average<-mean(feet/12+inches,na.rm=TRUE)
```
]]]

<br>

--

.font80[
- An exception to this rule are  `:`, `::` and `:::`.
]

.pull-left[ .code80[ .blockquote.exercise[ 

#### `r checker_green` Good
```{r}
x <- 1:10
base::get
```
]]]

.pull-right[ .code80[ .blockquote.exercise[ 

#### `r times_red` Bad

```{r}
x <- 1 : 10
base :: get
```
]]]

---
#### Spacing

.font80[
- Extra spacing (i.e. more than one space in a row) is ok if it improves alignment of equal signs or assignments (`<-`).
]

.code70[ .blockquote.exercise[ 

#### `r desktop` Example

```{r}
list(
  total = a + b + c, 
  mean  = (a + b + c) / n
)
```
]]

--

.font80[
- Do not place spaces around code in parentheses or square brackets (unless there’s a comma, in which case see above).
]

.pull-left[ .code70[ .blockquote.exercise[ 

#### `r checker_green` Good
```{r}
if (debug) do(x)
diamonds[5, ]
```
]]]

.pull-right[ .code70[ .blockquote.exercise[ 

#### `r times_red` Bad
```{r}
if ( debug ) do(x)  # No spaces around debug
x[1,]   # Needs a space after the comma
x[1 ,]  # Space goes after comma not before
```
]]]

---
#### Spacing

- An opening curly brace should never go on its own line and should always be followed by a new line. A closing curly brace should always go on its own line, unless it’s followed by else. Always indent the code inside curly braces.



.pull-left[ .blockquote.exercise[ 

#### `r checker_green`  Good

```{r}
if (y < 0 && debug) {
  message("Y is negative")
}

if (y == 0) {
  log(x)
} else {
  y ^ x
}
```
]]

--

.pull-right[ .blockquote.exercise[

#### `r times_red`  Bad
```{r}

if (y < 0 && debug) {
message("Y is negative") }

if (y == 0) {
  log(x)
} else {
  y ^ x
}
```
]]
---
#### Spacing

.font70[
- It’s okay to leave very short statements on the same line:
]

.code70[ .blockquote.exercise[

#### `r desktop` Example

```{r}
if (y < 0 && debug) message("Y is negative")
```
]]

.font70[
- Strive to limit your code to 80 characters per line. 
- If you find yourself running out of room, this is a good indication that you should encapsulate some of the work in a separate function.
]

--

#### Assignment

.font70[
- Use `<-`, not `=`, for assignment.]

.pull-left[.code70[ .blockquote.exercise[
#### `r checker_green` Good

```{r}
x <- 5
```
]]]

.pull-right[.code70[ .blockquote.exercise[
#### `r desktop` Bad

```{r}
x = 5
```
]]]

---
#### Commenting guidelines

- Comment your code. 
- Each line of a comment should begin with the comment symbol and a single space: `#`. 
- Comments should explain the why, not the what.
- Use commented lines of `-` and `=` to break up your file into easily readable chunks.

--

.blockquote.exercise[

#### `r desktop` Example

```{r}
# Load data ---------------------------

# Plot data ---------------------------
```
]

---
class: exercise_slide

## `r desktop` Exercises

<ol start = "1" >
<li> Combine all elements of <code>df_list</code> to one data frame. The result should be only a single line of code. </li>
</ol>
```{r}
df_list <- lapply(0:4, function(j) data.frame(c1 = (1:5) + 5 * j,
                                              c2 = letters[(1:5) + 5 * j]
                                              )
                  )
```

<ol start = "2" >
<li> Fix each of the following common data frame subsetting errors: </li>
</ol>
```{r}
mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]
```

<ol start = "3" >
<li>  What does <code>df[is.na(df)] <- 0</code> do? </li>
</ol>

<ol start = "4" >
<li>How would you randomly permute the columns of a data frame?</li>
</ol>

---
class: exercise_slide

## `r desktop` Exercises

<ol start = "5" >
<li>Write a <code>while</code> and a <code>repeat</code> loop doing the same as </li> 
</ol>
```{r}
for(i in LETTERS[1:10]){
  print(i)
}
```
<ol start = "6" >
<li> Write a <code>for</code> loop doing the same as</li>
</ol>

```{r}
count <- 0
repeat{
  x     <- sample(1:6, 1)
  count <- count + 1 
  if(x == 6) break
}
print(count)
```

---
class: exercise_slide

## `r desktop` Exercises


<ol start = "7" >
<li> .font90[What is the problem here? ] </li>
</ol>

.code80[
```{r}
# Number of apples
i <- 100
for (i in 1:3) {} 
paste("The number of apples is", i)
```
]
<ol start = "8" >
<li> .font90[What is the problem here? Can you debug it?] </li>
</ol>
.code80[
```{r}
df <- mtcars

lin_mod <- function() {
  lm(y ~ . ,data = df)
}

simple_lin_mod <- function(data, y, x) {
  df <- data[ ,c(y, x)]
  names(df) <- c("y", "x")
  fast_lm()
}

simple_lin_mod(df, "mpg", "cyl")

```
]