---
title: "Advanced R for Econometricians"
subtitle: "GenAI Applications with R"
author: 
  - "Martin Arnold"
  - "Martin Schmelzer"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false
    seal: false
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/navigate_home.js", "../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
---
class: title-slide title-ai center middle
```{r setup, echo = F, eval = T, warning = F}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE)
source(here::here("SoSe_2026/utils.R"))
```

```{r moon_reader, include=FALSE, eval = FALSE}
# setup for moon reader — copy to console for live preview
options(servr.interval = 0.5)
xaringan::inf_mr()
```

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`

---
## Chapter Overview

- Understanding what **Large Language Models (LLMs)** are
- Communicating with LLMs through **APIs**
- Setting up the **ellmer** package  
- Sending prompts and processing responses
- Using **system prompts** to shape LLM behavior
- Building multi-turn **conversations**
- Extracting **structured data** from text
- Building an **AI-powered study assistant** in R

---
## What Are Large Language Models?

A **Large Language Model (LLM)** is an AI system trained on vast amounts of text. Given a prompt, it generates a human-like response — think of it as an extremely well-read conversation partner.

.pull-left[
**What LLMs can do**

- Answer questions
- Summarize long texts
- Translate languages
- Generate creative writing
- Explain complex topics
- Help with coding
]

.pull-right[
**Popular LLMs**

.table-full[
| Model | Provider |
|-------|----------|
| GPT-4 | OpenAI |
| Claude | Anthropic |
| Llama | Meta |
| Gemini | Google |
]
]

You have likely used ChatGPT or similar tools in a web browser. Today we learn to control these models **programmatically** with R!

---
## How Programs Talk to LLMs: APIs

An **API** (Application Programming Interface) lets your R code send requests to a server and receive responses — just like a website, but designed for programs instead of humans.

.font90[
```{r api-diagram, echo=FALSE, eval=TRUE}
DiagrammeR::grViz('
digraph api_flow {
  graph [rankdir=LR, bgcolor=white, fontname="Arial", nodesep=1.0, ranksep=0.8]
  node [shape=box, style="filled,rounded", fontname="Arial", fontsize=16,
        penwidth=2, width=1.6, height=0.65]
  edge [fontname="Arial", fontsize=13, penwidth=1.8]

  R   [label="Your R Script", fillcolor="#004c93", fontcolor="white", color="#004c93"]
  API [label="API",           fillcolor="#dce8f5", fontcolor="#222",  color="#004c93"]
  LLM [label="LLM Server",   fillcolor="#dce8f5", fontcolor="#222",  color="#004c93"]

  R   -> API [label="  prompt + settings  ", color="#004c93", fontcolor="#333"]
  API -> LLM [label="  forward request  ",   color="#004c93", fontcolor="#333"]
  LLM -> R   [label="  generated text  ",    color="#888888", fontcolor="#333",
               style=dashed, constraint=false]
}
', height=250, width="90%")
```
]

Instead of typing into a chat window, your code sends a **prompt** to the API and gets back a **text response**. This means you can automate tasks, process data in bulk, and build custom tools!

---
## Why Groq?

**Groq** is an API provider that gives you access to powerful open-source LLMs (like Meta's Llama models) with a generous **free tier** — perfect for learning.

| Feature | Details |
|---------|---------|
| **Free API key** | Sign up at [console.groq.com](https://console.groq.com) |
| **Fast responses** | Groq's hardware is optimized for speed |
| **Model access** | Llama 3.3 70B, Llama 3.1 8B, and more |

We will use the **ellmer** package — a tidyverse-compatible R package that provides a unified interface to many LLM providers, including Groq.

.font90[
The skills you learn with Groq transfer directly to other providers. Switching from Groq to OpenAI or Google Gemini is a one-line change!
]

---
## The ellmer Package

**ellmer** is developed by Hadley Wickham and the Posit team. It makes it easy to use LLMs from R.

.pull-left[
**Key Features**

- Unified interface for 15+ providers
- Stateful conversation objects (R6)
- Streaming output
- Structured data extraction
- Tool/function calling
]

.pull-right[
**Supported Providers**

.font80[
| Function | Provider |
|----------|----------|
| `chat_groq()` | Groq |
| `chat_openai()` | OpenAI |
| `chat_anthropic()` | Anthropic |
| `chat_google_gemini()` | Google |
| `chat_ollama()` | Ollama (local) |
| `chat_github()` | GitHub Models |
]
]

All providers share the **same interface** — learn the pattern once, use it everywhere.

![:source_url https://ellmer.tidyverse.org](https://ellmer.tidyverse.org)

---
## Setting Up: Installation

Install the ellmer package from CRAN:

```{r}
install.packages("ellmer")
```

That is all you need. No additional packages for environment variables or HTTP — ellmer handles everything.

.font90[
| Package | Purpose |
|---------|---------|
| `ellmer` | Unified LLM client for R (includes Groq, OpenAI, Anthropic, ...) |
]

ellmer handles HTTP communication, message formatting, and response parsing — no additional packages required.

---
## Setting Up: Your API Key

1. Visit [console.groq.com](https://console.groq.com) and create a free account
2. Go to **API Keys** and generate a new key
3. Store the key in your `.Renviron` file. The easiest way: `usethis::edit_r_environ()`

    This opens your `.Renviron` file. Add this line and save:

    ```
    GROQ_API_KEY=gsk_your_key_here
    ```

4. Restart R (Ctrl/Cmd + Shift + F10) so the variable is loaded.
5. Verify it works:

    ```r
    # Check that the key is available (don't print the full key!)
    key <- Sys.getenv("GROQ_API_KEY")
    cat(substr(key, 1, 10), "...\n")
    ```

---
## Setting Up: Security

.content-box-white[

**Never put API keys directly in your code!**

Anyone who sees your code (e.g., on GitHub) could use and abuse your key. The `.Renviron` file keeps secrets safe because it lives outside your project files.

]

.font90[
**Best practices:**

- Store keys in `.Renviron` — not in your R scripts
- Add `.Renviron` to your `.gitignore` file
- Never print or `cat()` the full key

**Why `.Renviron`?**

- R loads `.Renviron` automatically on startup — no extra packages needed
- `Sys.getenv()` reads any environment variable
- `usethis::edit_r_environ()` makes editing easy
- ellmer reads the key automatically from the environment
]

---
## Your First API Call

Let's send a message to an LLM and get a response:

```{r}
library(ellmer)

chat <- chat_groq(                                              #<<
  system_prompt = "You are a helpful assistant."                #<<
)                                                               #<<

chat$chat("What is R in one sentence?")
```

```
## R is a programming language and software environment for statistical
## computing and graphics, widely used among statisticians and data miners
## for data analysis and developing statistical software.
```

Let's break this down step by step on the next slides.

---
## Breaking It Down

**Step 1: Create a Chat object**

```{r}
chat <- chat_groq(
  system_prompt = "You are a helpful assistant."
)
```

The **Chat object** connects to Groq's servers and stores the entire conversation history.

**Step 2: Send a message**

```{r}
chat$chat("What is R in one sentence?")
```

The `$chat()` method sends your prompt and streams the LLM's response to the console.

**Step 3: That's it!** No need to build message lists or parse JSON — ellmer handles everything.

.font80[
.blockquote.exercise[
#### `r ICONS$desktop` Try it yourself

Change the question to something related to your studies. Try: `"What is the difference between correlation and causation?"`
]]

---
## The Chat Object

The **Chat** object is an **R6 object** — a stateful, mutable object. This is different from typical R objects.

```{r}
chat <- chat_groq(system_prompt = "You are a helpful tutor.")
```

```{r}
# Send a message — the response is printed AND stored
chat$chat("What is a data frame?")
```

```{r}
# The Chat object remembers everything
print(chat)
```

```
## <Chat Groq/llama-3.3-70b-versatile turns=3 tokens=23/92>
## ── system ────────────────────────────────────────────────
## You are a helpful tutor.
## ── user ──────────────────────────────────────────────────
## What is a data frame?
## ── assistant ─────────────────────────────────────────────
## A data frame is a two-dimensional data structure in R ...
```

Notice the summary: **turns**, **tokens used**, and the full conversation history. This is how ellmer keeps track of context.

---
## Understanding Tokens

**Tokens** are the units LLMs use to process text. On average, one English word requires about **1.5 tokens**.

```{r}
# See token usage for the current session
token_usage()
```

```
##   provider              model input output price
## 1     Groq llama-3.3-70b-versatile    23     92 $0.00
```

| Concept | Approximate Tokens |
|---------|-------------------|
| A word | ~1.5 tokens |
| A sentence | ~20 tokens |
| A page of text | ~375–400 tokens |
| A full book | ~75,000–150,000 tokens |

API providers typically charge per million tokens. Groq's free tier is generous for learning, but it is good practice to monitor usage.

---
## Disabling Streaming

By default, `$chat()` streams the response to the console as it arrives. For **programmatic use**, you can disable this:

```{r}
# Capture the response silently
chat <- chat_groq(
  system_prompt = "Be concise.",
  echo = "none"                       #<<
)
response <- chat$chat("What is GDP?")
```

```{r}
# response is a character string you can work with
cat(response)
nchar(response)
```

| `echo` Setting | Behavior |
|----------------|----------|
| `"output"` | Stream response to console (default in interactive use) |
| `"none"` | Return silently — best for scripts and functions |
| `"all"` | Echo both input and output |

---
## System Prompts

The **system prompt** sets the LLM's behavior and personality. Think of it as giving the model its "job description."

```{r}
chat <- chat_groq(
  system_prompt = "You are a statistics tutor for graduate students."  #<<
)
chat$chat("Explain heteroskedasticity.")
```

The system prompt is powerful — it shapes how the LLM responds to **everything** that follows. 

.font90[
.blockquote.exercise[
#### `r ICONS$desktop` Try it yourself

Create Chat objects with different system prompts and ask the same question:

```{r}
q <- "What is the Pythagorean theorem?"

chat1 <- chat_groq(system_prompt = "Explain like I am 5 years old.")
chat1$chat(q)

chat2 <- chat_groq(system_prompt = "Answer in exactly one sentence.")
chat2$chat(q)

chat3 <- chat_groq(system_prompt = "You are a pirate. Answer in pirate speak.")
chat3$chat(q)
```
]]

---
## Shaping Behavior with System Prompts

Let's write a helper function that makes it easy to try different system prompts:

```{r}
ask <- function(system_message, user_question) {
  chat <- chat_groq(
    system_prompt = system_message,
    echo = "none"
  )
  chat$chat(user_question)
}
```

```{r}
# Try different "personalities"
question <- "What is OLS regression?"

ask("Explain like I am 5 years old.", question)
ask("Answer in exactly one sentence.", question)
ask("You are a strict statistics professor. Be formal and precise.", question)
```

Notice the pattern: **same question, different system prompts, completely different responses.** This is the core of **prompt engineering**.

---
## Multi-Turn Conversations

Unlike a simple function call, the Chat object **remembers** the entire conversation. Each new message builds on the previous context.

```{r}
chat <- chat_groq(system_prompt = "You are a helpful R tutor.")

chat$chat("What is a factor in R?")
```

```{r}
# The follow-up uses context from the previous exchange
chat$chat("Can you give me an example?")                        #<<
```

```{r}
# And another follow-up
chat$chat("How is it different from a character vector?")       #<<
```

Each time you call `$chat()`, ellmer sends **all previous messages** to the API so the LLM understands the full context. The conversation grows automatically.

---
## Multi-Turn Conversations: Under the Hood

The Chat object manages the conversation history for you. Each call to `$chat()` **appends two new turns**: your message and the assistant's response.

.pull-left[
**After 1 turn**
.code70[
```{r}
chat <- chat_groq(
  system_prompt = "You are a tutor."
)
chat$chat("What is a factor?")
print(chat)
```
```
## <Chat Groq turns=3 tokens=15/87>
## ── system ─────────
## You are a tutor.
## ── user ───────────
## What is a factor?
## ── assistant ──────
## A factor is ...
```
]]

.pull-right[
**After 2 turns**
.code70[
```{r}
chat$chat("Give an example.")
print(chat)
```
```
## <Chat Groq turns=5 tokens=22/168>
## ── system ─────────
## You are a tutor.
## ── user ───────────
## What is a factor?
## ── assistant ──────
## A factor is ...
## ── user ───────────
## Give an example.
## ── assistant ──────
## Here is an example ...
```
]]

Each turn includes the full prior context — the LLM "remembers" all previous messages automatically.

---
## Inspecting the Conversation

You can inspect the conversation history at any time:

```{r}
# Print the Chat object to see the full history
print(chat)
```

```{r}
# Get all turns as a list
turns <- chat$get_turns()
length(turns)  # number of turns
```

```{r}
# Access individual turns
turns[[1]]  # system turn
turns[[2]]  # first user turn
turns[[3]]  # first assistant response
```

This is useful for logging, debugging, or post-processing the conversation.

---
## Interactive Chat

For an interactive, ChatGPT-like experience directly in your R console, use `live_console()`:

```{r}
chat <- chat_groq(system_prompt = "You are a friendly R tutor.")
live_console(chat)                                               #<<
```

```
## ╔══════════════════════════════════════════════════════════╗
## ║  Entering chat console. Use """ for multi-line input.   ║
## ║  Press Ctrl+C to quit.                                  ║
## ╚══════════════════════════════════════════════════════════╝
## >>> What is a tibble?
## A tibble is a modern version of a data frame in R ...
```

After exiting (Ctrl+C), the conversation history is preserved in the `chat` object — you can continue programmatically.

For a richer experience, `live_browser()` opens a chat interface in your **web browser**:

```{r}
chat <- chat_groq(system_prompt = "You are a data science assistant.")
live_browser(chat)  # requires shinychat: install.packages("shinychat")  #<<
```

---
## Controlling Output: Temperature

The **temperature** parameter controls how creative vs. predictable the LLM's responses are.

| Temperature | Behavior | Best For |
|-------------|----------|----------|
| `0.0` | Deterministic, focused | Factual answers, summaries |
| `0.5` | Balanced | General-purpose use |
| `1.0` | Creative, varied | Brainstorming, creative writing |
| `> 1.0` | Very random | Experimental, playful |

```{r}
# Temperature is set when creating the Chat object
# Low temperature — deterministic
chat <- chat_groq(echo = "none", params = params(temperature = 0))     #<<
chat$chat("Write a haiku about coding.")

# High temperature — creative, varied
chat <- chat_groq(echo = "none", params = params(temperature = 1.5))   #<<
chat$chat("Write a haiku about coding.")
```

.font80[
.blockquote.exercise[
#### `r ICONS$desktop` Try it yourself

Create two Chat objects with `temperature = 0` and `temperature = 1.5`. Ask both the same question twice — which temperature gives more consistent results? **Tip:** Start with `temperature = 0` for study tasks.
]]

---
## Switching Providers

One of ellmer's strengths: **switching providers is a one-line change.** The rest of your code stays identical.

```{r}
# Using Groq (free, fast)
chat <- chat_groq(system_prompt = "You are a helpful assistant.")      #<<

# Using OpenAI's GPT-4
chat <- chat_openai(system_prompt = "You are a helpful assistant.")    #<<

# Using Anthropic's Claude
chat <- chat_anthropic(system_prompt = "You are a helpful assistant.") #<<

# Using Google Gemini (generous free tier)
chat <- chat_google_gemini(system_prompt = "You are a helpful assistant.") #<<

# Using a local model via Ollama (free, private)
chat <- chat_ollama(system_prompt = "You are a helpful assistant.")    #<<
```

After creating the chat object, the rest is **identical** — `$chat()`, `$get_turns()`, `live_console()`, etc. Each provider needs its own API key in `.Renviron` (see `?chat_openai`).

---
## Use Case 1: Summarize Your Notes

University students deal with long readings and lecture notes. Let's build a text summarizer:

.code80[
```{r}
summarize_text <- function(text, length = "short") {
  system_msg <- paste(
    "You are a study assistant. Summarize the following text in a",
    length, "and clear way. Use bullet points for key ideas."
  )
  
  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(text)
}
```
]

```{r}
# Summarize text directly
summarize_text("Heteroskedasticity means the variance of the error
  term is not constant across observations. This violates one of the
  Gauss-Markov assumptions and makes OLS estimators inefficient...")
```

```{r}
# Or read from a file first
notes <- readLines("my_notes.txt") |> paste(collapse = "\n")
cat(summarize_text(notes))
```

---
## Use Case 2: Generate Flashcards

Flashcards are a proven study technique. Let's have the LLM create them automatically:

```{r}
generate_flashcards <- function(topic, count = 5) {
  system_msg <- paste0(
    "You are a study assistant. Generate exactly ", count,
    " flashcards for the given topic. ",
    "Format each as:\nQ: [question]\nA: [concise answer]\n"
  )
  
  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(paste("Create flashcards about:", topic))
}
```

```{r}
cat(generate_flashcards("R data structures: vectors, lists, and data frames"))
```

```
## Q: What is the key difference between an atomic vector and a list?
## A: Atomic vectors can only contain elements of a single type,
##    while lists can contain elements of different types.
## ...
```

---
## Use Case 3: Improve Your Writing

Whether it is an essay, lab report, or email — the LLM can suggest improvements:

```{r}
focus_instructions <- list(                                      #<<
  clarity  = "Make the text clearer and easier to understand.",
  academic = "Make the text more formal and academic in tone.",
  concise  = "Make the text shorter while keeping the key message.",
  grammar  = "Fix grammar and spelling errors only."
)
```

```{r}
improve_writing <- function(text, focus = "clarity") {
  instruction <- focus_instructions[[focus]]
  system_msg <- paste("You are a writing tutor.", instruction)
  
  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(text)
}
```

```{r}
draft <- "the experiment showed that like the thing we thought
  would happen did happen mostly"

cat(improve_writing(draft, focus = "academic"))
```

---
## A Reusable Helper Function

Notice the pattern? Every LLM task follows the **same structure**. Let's create a reusable helper:

```{r}
ask_llm <- function(user_message,
                    system_message = "You are a helpful assistant.",
                    temperature = 0.5) {
  chat <- chat_groq(
    system_prompt = system_message,
    echo = "none",
    params = params(temperature = temperature)
  )
  chat$chat(user_message)
}
```

Now building new tools becomes a one-liner:

```{r}
# Summarize (assuming long_text holds your notes)
ask_llm(long_text, system_message = "Summarize in 3 bullet points.")

# Translate
ask_llm("Hello world", system_message = "Translate to German.")

# Quiz me
ask_llm("R loops", system_message = "Ask me a quiz question about this topic.")
```

---
## Structured Data Extraction

One of ellmer's most powerful features: you can ask the LLM to return **structured R objects** instead of free text.

```{r}
chat <- chat_openai(echo = "none")  # Groq does not support structured data  #<<

result <- chat$chat_structured(
  "My name is Susan and I'm 13 years old",
  type = type_object(                                            #<<
    name = type_string(),                                        #<<
    age  = type_number()                                         #<<
  )                                                              #<<
)
```

```
## List of 2
##  $ name: chr "Susan"
##  $ age : num 13
```

Instead of parsing text, you get a **named list** — ready for analysis. You define the desired structure with `type_object()`, `type_string()`, `type_number()`, etc.

---
## Structured Data: Data Frames

Entire **data frames** can be extracted from unstructured text:

.code80[
```{r}
prompt <- "
  John Smith: Age 30. Income $55,000. Education: B.A.
  Jane Doe: Age 25. Income $42,000. Education: M.Sc.
  Jose Rodriguez: Age 40. Income $78,000. Education: Ph.D.
"

type_person <- type_object(
  name      = type_string("Full name"),
  age       = type_integer("Age in years"),
  income    = type_number("Annual income in USD"),
  education = type_enum(c("B.A.", "M.Sc.", "Ph.D."), "Highest degree")
)

chat <- chat_openai(echo = "none")
people <- chat$chat_structured(prompt, type = type_array(type_person))
```
]

```
## # A tibble: 3 × 4
##   name             age income education
##   <chr>          <int>  <dbl> <fct>    
## 1 John Smith        30  55000 B.A.     
## 2 Jane Doe          25  42000 M.Sc.    
## 3 Jose Rodriguez    40  78000 Ph.D.    
```

---
## Structured Data: Type Specifications

.font90[
The `type_*()` functions define what structure you expect from the LLM:

| Function | R Equivalent | Example |
|----------|-------------|---------|
| `type_string()` | `character(1)` | `"hello"` |
| `type_number()` | `numeric(1)` | `3.14` |
| `type_integer()` | `integer(1)` | `42L` |
| `type_boolean()` | `logical(1)` | `TRUE` |
| `type_enum(values)` | `factor` | One of `c("A", "B", "C")` |
| `type_array(item_type)` | `vector` or `list` | `c(1, 2, 3)` |
| `type_object(...)` | `named list` | `list(name = "X", age = 30)` |
]

**Key insight:** A `type_array()` of `type_object()`s becomes a **tibble** — this is how you extract tabular data from unstructured text.

.font80[
.blockquote.exercise[
#### `r ICONS$desktop` Try it yourself

Use `$chat_structured()` with `chat_openai()` or `chat_anthropic()` to extract the **sentiment** (positive/negative/neutral) and a **confidence score** from a product review. (Groq does not support structured extraction.)
]]

---
## Building the Study Assistant

Let's combine everything into a complete interactive program. We use our `ask_llm()` helper and R's `readline()` for user input:

.code80[
```{r}
show_menu <- function() {
  cat("\n=== Study Assistant ===\n")
  cat("1 - Summarize text\n")
  cat("2 - Generate flashcards\n")
  cat("3 - Improve my writing\n")
  cat("4 - Ask a question\n")
  cat("q - Quit\n")
  readline("Choose an option: ")
}
```
]

---
## Building the Study Assistant: Main Loop

The `repeat` loop calls `show_menu()` and dispatches to the appropriate helper:

.code60[
```{r}
# Assumes ask_llm() is defined as on the previous slide

repeat {
  choice <- show_menu()
  
  if (choice == "q") {
    cat("Happy studying!\n")
    break
  }
  
  switch(choice,
    "1" = {
      text <- readline("Enter text to summarize: ")
      cat("\nSummary:\n")
      cat(ask_llm(text, "Summarize in clear bullet points.", temperature = 0.3))
    },
    "2" = {
      topic <- readline("Enter a topic: ")
      count <- readline("How many flashcards? (default: 5) ")
      if (count == "") count <- "5"
      sys_msg <- paste0("Generate ", count, " flashcards. Format: Q: ... / A: ...")
      cat(ask_llm(topic, sys_msg, temperature = 0.5))
    },
    "3" = {
      text <- readline("Enter text to improve: ")
      style <- readline("Focus (clarity/academic/concise): ")
      if (style == "") style <- "clarity"
      sys_msg <- paste("Improve this text. Focus on", style, ". Show improved version.")
      cat(ask_llm(text, sys_msg, temperature = 0.3))
    },
    "4" = {
      question <- readline("Your question: ")
      cat(ask_llm(question))
    },
    cat("Invalid option. Try again.\n")
  )
}
```
]

---
## Building the Study Assistant (continued)

This program uses concepts from across the course:

| Concept | How It's Used |
|---------|---------------|
| **Functions** | `ask_llm()`, `show_menu()` |
| **Named lists** | `focus_instructions` |
| **Control flow** | `repeat` loop, `switch()`, `break` |
| **String manipulation** | `paste()`, `paste0()`, `cat()` |
| **R6 objects** | Chat objects inside `ask_llm()` |
| **API interaction** | `chat_groq()`, `$chat()` |

.font90[
.blockquote.exercise[
#### `r ICONS$desktop` Extension ideas

1. Add a "conversation mode" using `live_console()` 
2. Read text from a file with `readLines()` for option 1
3. Add a "translate" option
4. Log the conversation to a file
]]

---
## Error Handling for API Calls

API calls can fail — the network might be down, your key might be invalid, or you might hit rate limits. Use `tryCatch()`:

.code80[
```{r}
ask_llm_safe <- function(user_message,
                         system_message = "You are a helpful assistant.") {
  tryCatch(
    {
      chat <- chat_groq(system_prompt = system_message, echo = "none")
      chat$chat(user_message)
    },
    error = function(e) {
      msg <- conditionMessage(e)
      if (grepl("401|authentication|unauthorized", msg, ignore.case = TRUE)) {
        "Error: Invalid API key. Check your .Renviron file."
      } else if (grepl("429|rate.limit", msg, ignore.case = TRUE)) {
        "Error: Too many requests. Wait a moment and try again."
      } else if (grepl("connection|network|timeout", msg, ignore.case = TRUE)) {
        "Error: Cannot reach the API. Check your internet connection."
      } else {
        paste("Error:", msg)
      }
    }
  )
}
```
]

Always wrap API calls in `tryCatch()` for production code. The error messages help users diagnose the problem.

---
## Best Practices

.pull-left[
**Prompt Engineering**

| Tip | Example |
|-----|---------|
| Be specific | "List 5 key points" not "Summarize" |
| Set the format | "Answer in bullet points" |
| Give examples | "Format like: Q: ... A: ..." |
| Set constraints | "Use max 100 words" |
]

.pull-right[
**Security & Cost**

| Tip | Why |
|-----|-----|
| Use `.Renviron` | Keep keys out of code |
| Add to `.gitignore` | Prevent accidental sharing |
| Monitor `token_usage()` | Free tiers have limits |
| Use low temperature | More focused output |
]

**Prompt engineering** — the art of writing effective instructions for LLMs — is a skill in itself. The better your system prompt, the better the results. Experiment!

---
## Local LLMs with Ollama

For privacy-sensitive data — or to avoid API costs — run an LLM on your own machine with **Ollama**:

1. Install Ollama from [ollama.com](https://ollama.com) — free, runs locally
2. Pull a model: run `ollama pull llama3.2` in your terminal
3. Use it from R via `chat_ollama()` — the interface is **identical** to cloud providers:

```{r}
# Exactly the same workflow — just a different provider function
chat <- chat_ollama(
  model = "llama3.2",
  system_prompt = "You are a helpful assistant."
)
chat$chat("What is heteroskedasticity?")
```

.pull-left[
**Cloud (Groq, OpenAI, ...)**
- More powerful models
- Fast inference
- Requires API key
- Data leaves your machine
]

.pull-right[
**Local via Ollama**
- Completely private
- Always free to run
- Requires decent hardware
- Same `ellmer` interface
]

---
## Summary

.font90[

.pull-left[
| Topic | Key Points |
|-------|------------|
| **LLMs** | AI models that generate text |
| **APIs** | Send requests, get responses |
| **ellmer** | `install.packages("ellmer")` |
| **Chat object** | `chat_groq(system_prompt = ...)` |
| **Send message** | `chat$chat("question")` |
]

.pull-right[
| Topic | Key Points |
|-------|------------|
| **System prompt** | Shapes LLM behavior |
| **Conversation** | Stateful R6 object |
| **Temperature** | 0 = focused, 1+ = creative |
| **Structured data** | `$chat_structured(type = ...)` |
| **Error handling** | `tryCatch()` for API calls |
]
]

<div style="clear:both"></div>

You built an **AI-powered study assistant** using R concepts: functions, named lists, control flow, string manipulation, R6 objects, and error handling!

.font90[
**Further reading:**
- [ellmer documentation](https://ellmer.tidyverse.org)
- [Prompt design vignette](https://ellmer.tidyverse.org/articles/prompt-design.html)
- [Structured data vignette](https://ellmer.tidyverse.org/articles/structured-data.html)
]

---
class: exercise_slide

## `r ICONS$desktop` Exercises

.font90[
<ol>
<li> <b>Basic chat:</b> Create a <code>chat_groq()</code> object with a system prompt that makes the LLM act as an econometrics tutor. Ask it to explain the difference between fixed and random effects. </li>

<li> <b>System prompt comparison:</b> Write a function <code>explain(topic, audience)</code> that takes a topic and an audience description (e.g., "a 5 year old", "a PhD student") and returns an explanation tailored to that audience. Test it with "instrumental variables." </li>

<li> <b>Summarizer:</b> Read a text file of your choice with <code>readLines()</code> and use <code>ask_llm()</code> to produce a bullet-point summary. Experiment with different temperature values (0, 0.5, 1.0) — how do the summaries differ? </li>

<li> <b>Multi-turn conversation:</b> Build a conversation where you first ask "What is endogeneity?", then follow up with "Give me an example from labor economics", then "How would I test for it?". Print the Chat object to see the full history. </li>

<li> <b>Structured extraction:</b> Use <code>chat_openai()</code> (not Groq) and <code>$chat_structured()</code> to extract country name, GDP (numeric), and continent from this text: <i>"Germany has a GDP of approximately 4.2 trillion USD and is located in Europe. Brazil's GDP is about 1.9 trillion USD, in South America."</i> The result should be a tibble. </li>

<li> <b>Provider comparison:</b> Use <code>chat_groq()</code> and <code>chat_google_gemini()</code> (or another provider) to answer the same question. Compare the responses and token usage with <code>token_usage()</code>. </li>
</ol>
]
