---
title: "Advanced R for Econometricians"
subtitle: "Building GenAI Applications with R"
author: 
  - "Martin Arnold"
  - "Martin Schmelzer"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false
    seal: false
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/navigate_home.js", "../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
---
class: title-slide title-ai center middle
```{r setup, echo = F, eval = T, warning = F}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE)
source(here::here("SoSe_2026/utils.R"))
```

```{r moon_reader, include=FALSE, eval = FALSE}
# setup for moon reader — copy to console for live preview
options(servr.interval = 0.5)
xaringan::inf_mr()
```

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`

???
Workshop target: 4 to 4.5 hours including setup friction, live coding, Q&A, and joint exercises.

---
## Today: Build Small AI Tools in R

.pull-left[
**By the end, you can**

- call LLM APIs from R
- control behavior with prompts
- keep multi-turn context
- send document scans to a vision model
- extract structured data
- wrap LLM calls in reusable functions
]

.pull-right[
**Workshop rhythm**

- explain the idea
- live-code the core pattern
- solve a short exercise
- discuss failure modes
]

???
Emphasize that the deck is intentionally not a cookbook on the visible slides. The code appears in presenter notes because the goal is to build it together.

---
## The Core Mental Model

The programming skill is not "chatting". It is turning repeated language tasks into reliable R workflows.

An LLM call essentially is a _remote_ function call:

<br/>

| Part | Role |
|------|------|
| **model** | the engine that generates the answer |
| **system prompt** | persistent instruction for behavior |
| **user message** | the task-specific input |
| **parameters** | settings such as temperature |
| **response** | text or structured data returned to R |

---
## API Call – How R Talks to an LLM
#### Schema for a single LLM request

.font90[
```{r api-diagram, echo=FALSE, eval=TRUE}
DiagrammeR::grViz('
digraph api_flow {
  graph [rankdir=LR, bgcolor=white, fontname="Arial", nodesep=1.0, ranksep=0.8]
  node [shape=box, style="filled,rounded", fontname="Arial", fontsize=16,
        penwidth=2, width=1.8, height=0.65]
  edge [fontname="Arial", fontsize=13, penwidth=1.8]

  R   [label="R", fillcolor="#004c93", fontcolor="white", color="#004c93"]
  API [label="Provider API", fillcolor="#dce8f5", fontcolor="#222", color="#004c93"]
  LLM [label="LLM", fillcolor="#dce8f5", fontcolor="#222", color="#004c93"]

  R   -> API [label=" prompt + inputs ", color="#004c93", fontcolor="#333"]
  API -> LLM [label=" request ", color="#004c93", fontcolor="#333"]
  LLM -> API [label=" response ", color="#888888", fontcolor="#333", style=dashed]
  API -> R   [label=" response ", color="#888888", fontcolor="#333", style=dashed]
}
', height=240, width="90%")
```
]

The R session sends the request. The provider runs the model. R receives the result as an object we can print, save, parse, or pass to another function.

The API is **programming language agnostic**: it communicates via standardized HTTP requests and JSON-formatted messages. Python, Julia, or any other language can use the same API — only the client library differs.

---
## Why Groq + ellmer?

.pull-left[
**Groq**

- fast hosted inference
- free API key for getting started
- open-weight models
- text and vision models
]

.pull-right[
**ellmer**

- one interface for many providers
- stateful chat objects
- streaming output
- image inputs
- structured extraction
]

Example Groq model strings: `llama-3.3-70b-versatile`, `llama-3.1-8b-instant`, `meta-llama/llama-4-scout-17b-16e-instruct`.

![:source_url https://ellmer.tidyverse.org](https://ellmer.tidyverse.org)

---
## Setup Checklist

.font90[
| Need | Action |
|------|--------|
| R package | `install.packages("ellmer")` |
| Groq key | create one at [console.groq.com](https://console.groq.com) |
| Environment variable | add `GROQ_API_KEY=...` to user or project `.Renviron` |
| Restart | restart R so the key is loaded |
| Browser chat later | install `shinychat` if using `live_browser()` |
]

<br>
.content-box-white[
Never put API keys directly in scripts, slides, or GitHub repositories.
]

???
Live setup check:

```r
install.packages("ellmer")
usethis::edit_r_environ()

key <- Sys.getenv("GROQ_API_KEY")
nzchar(key)
substr(key, 1, 8)
```

If many students are still setting up, have them pair with someone whose key works.

---
## The Minimal Pattern

.font90[
A basic call to the LLM requires the following:

1. Load `ellmer`
2. Create a `chat_*()` object
3. Send a message with the `$chat()` _method_
4. Work with the returned character string
]

???
Live code:

```r
library(ellmer)

chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are a helpful assistant."
)

chat$chat("What is R in one sentence?")
```

Discussion prompts:
- What is hard-coded?
- What could become a function argument?
- What will break if the API key is missing?

---
class: exercise_slide

## `r ICONS$desktop` Exercise: One Prompt, Three Terms

.font90[
Write a **single system prompt** that forces the model to always respond with:

- exactly one sentence defining the term
- one concrete empirical example

Then send three econometric terms as separate user messages — **without creating a new chat object each time**.

Suggested terms: *heteroskedasticity*, *endogeneity*, *multicollinearity*.
]

<br>
.content-box-white[
Goal: the system prompt is a persistent contract. The user message is the only thing that changes between calls.
]

???
Joint solution:

```r
library(ellmer)

chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = paste(
    "You are a concise econometrics tutor.",
    "For every term the user sends, respond with exactly two lines:",
    "Line 1: a one-sentence definition.",
    "Line 2: one concrete empirical example."
  ),
  echo = "none"
)

chat$chat("heteroskedasticity")
chat$chat("endogeneity")
chat$chat("multicollinearity")
```

Discussion:
- What happens to the format across all three responses?
- What would break if you put the format instruction in the user message instead?
- How would you change the system prompt to get LaTeX-formatted definitions?

---
## The Chat Object Remembers

.pull-left[
**What it stores**

- user turns
- assistant turns
- token usage
- cost information
- model/provider metadata
]

.pull-right[
**Why it matters**

- follow-up questions work
- debugging becomes easier
- context can grow expensive
- results depend on history
]

"Memory" is a feature of `groq_chat()`: Using "Bare" API calls to an LLM, we'd have to implement this ourselves! 

The system prompt is stored separately, but it is sent _everytime_ with the updated conversation (which is also sent along).

???
Live code:

```r
chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are a helpful R tutor."
)

chat$chat("What is a data frame?")
print(chat)
```

Point out `turns`, `input`, `output`, and `cost` in the printed object.

Token usage: LLMs process tokens, not words (roughly 1 word = 1–2 tokens, 1 page ≈ 400 tokens). Call `token_usage()` to inspect session totals. Costs grow fast when looping over many documents.

---
## Streaming vs. Capturing Output

.pull-left[
**Interactive use**

Let the answer stream to the console.
]

.pull-right[
**Programmatic use**

Set `echo = "none"` and assign the result.
]

<br>
.content-box-white[
For functions and scripts, capturing the response is usually the safer default.
]

???
Live code:

```r
chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "Be concise.",
  echo = "none"
)

response <- chat$chat("What is GDP?")
cat(response)
nchar(response)
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Capture and Inspect

.font90[
Write code that:

1. asks for a one-sentence definition of **heteroskedasticity**
2. stores the result in `answer`
3. prints the answer
4. counts the characters with `nchar()`
]

<br>
.content-box-white[
This is the first step toward turning prompts into reusable functions.
]

???
Joint solution:

```r
chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are a precise econometrics tutor.",
  echo = "none"
)

answer <- chat$chat("Define heteroskedasticity in one sentence.")
cat(answer)
nchar(answer)
```

Possible follow-up:
- Add "Do not use more than 30 words" to the prompt.

---
## System Prompts Shape Behavior

.font90[
The system prompt tells the model how to behave across the conversation.

Good system prompts often specify:

- role
- audience
- style
- constraints
- output format
]

<br>
.content-box-white[
Small prompt changes can cause large response changes. Test prompts with the same user question.
]

???
Live code:

```r
q <- "What is OLS regression?"

chat1 <- chat_groq(
  system_prompt = "Explain like I am 5 years old.",
  echo = "none"
)
chat1$chat(q)

chat2 <- chat_groq(
  system_prompt = "Answer in exactly one sentence.",
  echo = "none"
)
chat2$chat(q)

chat3 <- chat_groq(
  system_prompt = "You are a strict econometrics professor. Be formal and precise.",
  echo = "none"
)
chat3$chat(q)
```

---
## Make Prompt Experiments Reusable

.font90[
Instead of manually creating a new chat object each time, we may write a with the following interface:

<br/>

| Parameter | Meaning |
|-------|---------|
| `system_message` | role and constraints |
| `user_question` | task |
| `return` value | model response |
]

???
Live code:

```r
ask <- function(system_message, user_question) {
  chat <- chat_groq(
    model = "llama-3.3-70b-versatile",
    system_prompt = system_message,
    echo = "none"
  )
  chat$chat(user_question)
}

question <- "What is OLS regression?"

ask("Explain like I am 5 years old.", question)
ask("Answer in exactly one sentence.", question)
ask("You are a strict econometrics professor. Be formal and precise.", question)
```

---
## Multi-Turn Conversations

.pull-left[
**One-off call**

- independent task
- easier to reproduce
- cheaper context
]

.pull-right[
**Conversation**

- follow-ups use context
- useful for tutoring
- can drift over time
]

.font90[
Each new chat message appends to the conversation. ellmer sends the relevant history with later requests. Use `get_turns(include_system_prompt = TRUE)` to inspect or log the accumulated turns.
]

???
Live code:

```r
chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are a helpful R tutor."
)

chat$chat("What is a factor in R?")
chat$chat("Can you give me an example?")
chat$chat("How is it different from a character vector?")

print(chat)
```

Inspect turns:

```r
turns <- chat$get_turns(include_system_prompt = TRUE)
length(turns)
turns[[1]]
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Context Matters

.font90[
Build a conversation:

1. "What is endogeneity?"
2. "Give me an example from labor economics."
3. "How would I test for it?"

Then print the chat object and inspect the turns.
]

<br>
.content-box-white[
The second and third questions are intentionally vague. They rely on context.
]

???
Joint solution:

```r
chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are an econometrics tutor. Be concise."
)

chat$chat("What is endogeneity?")
chat$chat("Give me an example from labor economics.")
chat$chat("How would I test for it?")

print(chat)
turns <- chat$get_turns(include_system_prompt = TRUE)
length(turns)
```

Discussion:
- Would this be better as one prompt or a conversation?
- What might happen after 20 turns?

---
## Interactive Chat

.font90[
ellmer also offers interactive modes:

| Function | Interface |
|----------|-----------|
| `live_console()` | ChatGPT-like console session |
| `live_browser()` | browser chat interface |
]

<br>
.content-box-white[
Use these for exploration. Use `$chat()` inside functions and scripts.
]

???
Live code:

```r
chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are a friendly R tutor."
)

live_console(chat)
```

Browser version:

```r
install.packages("shinychat")

chat <- chat_groq(
  model = "llama-3.3-70b-versatile",
  system_prompt = "You are a data science assistant."
)

live_browser(chat)
```

---
## Multimodal Models

.font90[
Some models accept **text plus images**.

This is useful for:

- scanned forms
- screenshots
- handwritten notes
- plots and figures
- tables embedded in PDFs or images
]

<br>
.content-box-white[
For Groq vision input, use a multimodal model such as `meta-llama/llama-4-scout-17b-16e-instruct`.
]

---
## Document Scan to Markdown

.font90[
The workflow:

1. choose a vision-capable model
2. attach the local scan with `content_image_file()`
3. ask for Markdown only
4. review the output before using it
]

<br>
.content-box-white[
This is OCR plus document reconstruction. Treat the result as a draft, not ground truth.
]

???
Live code:

```r
library(ellmer)

# https://www.reddit.com/r/interestingasfuck/comments/srertt/spanish_healthcare_medicine_invoice/

cwd <- getwd()
document_path <- paste0(cwd, "/SoSe_2026/AI/med_invoice.webp")  # use your own JPG or PNG scan
chat <- chat_groq(
  model = "meta-llama/llama-4-scout-17b-16e-instruct",
  system_prompt = paste(
    "You are a careful OCR assistant.",
    "Extract the visible document content and return only Markdown.",
    "Preserve headings, paragraphs, lists, tables, and signatures.",
    "Mark unreadable text as [illegible]."
  ),
  echo = "none",
  params = params(temperature = 0)
)
markdown_doc <- chat$chat(
  "Convert this scanned document into clean Markdown and translate to English.",
  content_image_file(document_path, resize = "high")
)
cat(markdown_doc)

```

If `magick` is not installed, either install it or use `resize = "none"` for already appropriately sized images.

Example output for a scanned registration form:

```markdown
# Application for Course Registration
## Student
- Name: Susan Miller, Student ID: 1234567, Program: Economics M.Sc.
## Requested Course
| Course | Semester | Credits |
|--------|----------|---------|
| Advanced R | Summer 2026 | 6 |
## Signature
Susan Miller, 2026-04-15
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Prompt the OCR

.font90[
Starting from the document-scan workflow, adapt the prompt so the model:

- returns only Markdown
- preserves tables
- marks uncertain text with `[unclear]`
- adds a final section called `## Extraction Notes`
]

<br>
.content-box-white[
This is prompt engineering for a concrete workflow, not a new API pattern.
]

???
Joint solution:

```r
ocr_prompt <- paste(
  "You are a careful OCR assistant.",
  "Return only Markdown.",
  "Preserve headings, paragraphs, lists, and tables.",
  "Mark uncertain text as [unclear].",
  "End with a section named ## Extraction Notes that briefly lists any quality issues."
)

chat <- chat_groq(
  model = "meta-llama/llama-4-scout-17b-16e-instruct",
  system_prompt = ocr_prompt,
  echo = "none",
  params = params(temperature = 0)
)
```

If no shared scan is available, write and compare the prompt only.

---
class: segue-red middle center

# 4. Parameters and Providers

---
## Temperature

.font90[
Temperature controls the randomness of the generated output. It hence determines the balance between predictability (low temperature) and creativity (high temperature) of the LLM.  

<br/>

| Temperature | Behavior | Good for |
|-------------|----------|----------|
| `0.0` | most predictable | extraction, factual summaries |
| `0.5` | balanced | general tutoring |
| `1.0` | more varied | brainstorming |
| `> 1.0` | very varied | experiments |
]

<br>
.content-box-white[
For study tools and extraction, start low.
]

???
Live code:

```r
low <- chat_groq(
  echo = "none",
  params = params(temperature = 0)
)
low$chat("Write a haiku about coding.")
low$chat("Write a haiku about coding.")

high <- chat_groq(
  echo = "none",
  params = params(temperature = 1.2)
)
high$chat("Write a haiku about coding.")
high$chat("Write a haiku about coding.")
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Stable Study Output

.font90[
Ask the same model to create **three flashcards about omitted variable bias**.

Run it twice with:

- `temperature = 0`
- `temperature = 1`

Which setting would you use for a study assistant?
]

???
Joint solution:

```r
prompt <- "Create exactly three flashcards about omitted variable bias. Format as Q: ... A: ..."

chat_low <- chat_groq(
  echo = "none",
  params = params(temperature = 0)
)

chat_high <- chat_groq(
  echo = "none",
  params = params(temperature = 1)
)

cat(chat_low$chat(prompt))
cat(chat_low$chat(prompt))

cat(chat_high$chat(prompt))
cat(chat_high$chat(prompt))
```

---
## Switching Providers

.font90[
The provider function changes; the chat interface stays the same.

<br/>

| Provider | Function |
|----------|----------|
| Groq | `chat_groq()` |
| OpenAI | `chat_openai()` |
| Anthropic | `chat_anthropic()` |
| Google Gemini | `chat_google_gemini()` |
| Ollama | `chat_ollama()` |
]

Each provider needs its own key or local setup.

???
Live code:

```r
chat <- chat_groq(system_prompt = "You are a helpful assistant.")
chat <- chat_openai(system_prompt = "You are a helpful assistant.")
chat <- chat_anthropic(system_prompt = "You are a helpful assistant.")
chat <- chat_google_gemini(system_prompt = "You are a helpful assistant.")

chat <- chat_ollama(
  model = "llama3.2",
  system_prompt = "You are a helpful assistant."
)
```

---
class: segue-red middle center

# 5. Reusable Study Tools

---
## Summarizer Design

.pull-left[
**Function arguments**

- `text`
- `detail`
- `audience`
]

.pull-right[
**Prompt requirements**

- preserve key ideas
- use bullet points
- avoid invented facts
- match the audience
]

???
Live code:

```r
summarize_text <- function(text, detail = "short", audience = "students") {
  system_msg <- paste(
    "You are a study assistant.",
    "Summarize the user's text for", audience, ".",
    "Use a", detail, "summary in bullet points.",
    "Do not add facts that are not in the text."
  )

  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(text)
}

txt <- paste(
  "Heteroskedasticity means the variance of the error term is not constant",
  "across observations. This violates one of the Gauss-Markov assumptions",
  "and makes the usual OLS standard errors unreliable."
)

cat(summarize_text(txt))
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Add a Constraint

.font90[
Extend `summarize_text()` so it also accepts `max_bullets`.

Then call it with:

- `detail = "concise"`
- `max_bullets = 3`
- a short paragraph from your notes
]

<br>
.content-box-white[
Practice turning prompt requirements into function arguments.
]

???
Joint solution:

```r
summarize_text <- function(text,
                           detail = "short",
                           audience = "students",
                           max_bullets = 3) {
  system_msg <- paste(
    "You are a study assistant.",
    "Summarize the user's text for", audience, ".",
    "Use a", detail, "summary.",
    "Use at most", max_bullets, "bullet points.",
    "Do not add facts that are not in the text."
  )

  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(text)
}
```

---
## Flashcards and Writing Feedback

.font90[
These tools differ mostly in the system prompt.

| Tool | Key constraint |
|------|----------------|
| flashcards | exact count and stable format |
| writing feedback | focus on one revision goal |
| translation | specify target language and preserve meaning |
]

<br>
.content-box-white[
A good prompt makes downstream parsing easier.
]

???
Live code:

```r
generate_flashcards <- function(topic, count = 5) {
  system_msg <- paste0(
    "You are a study assistant. Generate exactly ", count,
    " flashcards for the given topic. ",
    "Format each as:\nQ: [question]\nA: [concise answer]\n"
  )

  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(paste("Create flashcards about:", topic))
}

cat(generate_flashcards("R data structures: vectors, lists, and data frames"))
```

Writing helper:

```r
focus_instructions <- list(
  clarity  = "Make the text clearer and easier to understand.",
  academic = "Make the text more formal and academic in tone.",
  concise  = "Make the text shorter while keeping the key message.",
  grammar  = "Fix grammar and spelling errors only."
)

improve_writing <- function(text, focus = "clarity") {
  focus <- match.arg(focus, names(focus_instructions))
  instruction <- focus_instructions[[focus]]
  system_msg <- paste("You are a writing tutor.", instruction)

  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(text)
}
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Build a Translator

.font90[
Write `translate_text(text, target_language)` using the same pattern.

Requirements:

- preserve the meaning
- keep technical terms precise
- return only the translation
]

<br>
.content-box-white[
This is an instructive variation: new task, same API pattern.
]

???
Joint solution:

```r
translate_text <- function(text, target_language = "German") {
  system_msg <- paste(
    "You are a precise academic translator.",
    "Translate the user's text to", target_language, ".",
    "Preserve technical terms.",
    "Return only the translation."
  )

  chat <- chat_groq(system_prompt = system_msg, echo = "none")
  chat$chat(text)
}

cat(translate_text("Ordinary least squares estimates a linear conditional mean."))
```

---
## One General Helper

.font90[
For small tools, a single helper removes repetition:

| Argument | Purpose |
|----------|---------|
| `user_message` | task input |
| `system_message` | role and output contract |
| `temperature` | randomness control |
]

<br>
.content-box-white[
Use specialized functions when the task has its own validation or post-processing.
]

???
Live code:

```r
ask_llm <- function(user_message,
                    system_message = "You are a helpful assistant.",
                    temperature = 0.5) {
  chat <- chat_groq(
    system_prompt = system_message,
    echo = "none",
    params = params(temperature = temperature)
  )
  chat$chat(user_message)
}

ask_llm(
  "R loops",
  system_message = "Ask me one quiz question about this topic."
)
```

---
class: segue-red middle center

# 6. Structured Extraction

---
## When Text Is Not Enough

.font90[
Free text is easy to read, but hard to analyze.

Structured extraction asks the model to return an R object with a specified schema.

Useful for:

- names and dates
- classifications
- sentiment labels
- tables from text
- form-like records
]

<br>
.content-box-white[
Use a provider/model that supports structured data. In this deck we use OpenAI for these examples.
]

???
Live code — start with a simple two-field schema:

```r
chat <- chat_openai(echo = "none")

result <- chat$chat_structured(
  "My name is Susan and I'm 13 years old",
  type = type_object(
    name = type_string(),
    age  = type_number()
  )
)

str(result)  # returns a named list
```

The `type_object()` call defines the expected shape. The result is a named list, not prose.

---
## Extract a Data Frame

.font90[
A `type_array()` of `type_object()`s becomes tabular data.

Use this when the input text contains repeated records.
]

???
Live code:

```r
prompt <- "
  John Smith: Age 30. Income $55,000. Education: B.A.
  Jane Doe: Age 25. Income $42,000. Education: M.Sc.
  Jose Rodriguez: Age 40. Income $78,000. Education: Ph.D.
"

type_person <- type_object(
  name      = type_string("Full name"),
  age       = type_integer("Age in years"),
  income    = type_number("Annual income in USD"),
  education = type_enum(c("B.A.", "M.Sc.", "Ph.D."), "Highest degree")
)

chat <- chat_openai(echo = "none")
people <- chat$chat_structured(prompt, type = type_array(type_person))
people
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Sentiment Schema

.font90[
Extract from a product review:

- `sentiment`: one of `positive`, `negative`, `neutral`
- `confidence`: number between 0 and 1
- `reason`: one short sentence
]

<br>
.content-box-white[
This repeats the schema idea but adds classification and a confidence score.
]

???
Joint solution:

```r
review <- paste(
  "The app is easy to use and the charts are beautiful,",
  "but it crashes whenever I import a large CSV file."
)

type_review <- type_object(
  sentiment = type_enum(c("positive", "negative", "neutral")),
  confidence = type_number("Confidence between 0 and 1"),
  reason = type_string("One short sentence explaining the label")
)

chat <- chat_openai(echo = "none")
chat$chat_structured(review, type = type_review)
```

Discussion:
- Is mixed sentiment better represented by one label or multiple fields?

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Country Table

.font90[
Extract a tibble with:

- country
- GDP in trillion USD
- continent

from this text:

> Germany has a GDP of approximately 4.2 trillion USD and is located in Europe. Brazil's GDP is about 1.9 trillion USD, in South America.
]

???
Joint solution:

```r
country_text <- paste(
  "Germany has a GDP of approximately 4.2 trillion USD and is located in Europe.",
  "Brazil's GDP is about 1.9 trillion USD, in South America."
)

type_country <- type_object(
  country = type_string(),
  gdp_trillion_usd = type_number(),
  continent = type_string()
)

chat <- chat_openai(echo = "none")
countries <- chat$chat_structured(country_text, type = type_array(type_country))
countries
```

---
class: segue-red middle center

# 7. Assemble an App-Like Workflow

---
## Study Assistant: Architecture

.font90[
The interactive assistant combines ideas we already used:

| Component | R concept |
|-----------|-----------|
| menu | `readline()` |
| task selection | `switch()` |
| AI calls | `ask_llm()` |
| repeated interaction | `repeat` loop |
| exit | `break` |
]

<br>
.content-box-white[
The main slide shows the architecture. The complete code is better live-coded in pieces.
]

???
Live code:

```r
show_menu <- function() {
  cat("\n=== Study Assistant ===\n")
  cat("1 - Summarize text\n")
  cat("2 - Generate flashcards\n")
  cat("3 - Improve my writing\n")
  cat("4 - Ask a question\n")
  cat("q - Quit\n")
  readline("Choose an option: ")
}
```

Then build the loop:

```r
repeat {
  choice <- show_menu()

  if (choice == "q") {
    cat("Happy studying!\n")
    break
  }

  switch(choice,
    "1" = {
      text <- readline("Enter text to summarize: ")
      cat("\nSummary:\n")
      cat(ask_llm(text, "Summarize in clear bullet points.", temperature = 0.3))
    },
    "2" = {
      topic <- readline("Enter a topic: ")
      count <- readline("How many flashcards? (default: 5) ")
      if (count == "") count <- "5"
      sys_msg <- paste0("Generate ", count, " flashcards. Format: Q: ... / A: ...")
      cat(ask_llm(topic, sys_msg, temperature = 0.5))
    },
    "3" = {
      text <- readline("Enter text to improve: ")
      style <- readline("Focus (clarity/academic/concise/grammar): ")
      if (style == "") style <- "clarity"
      sys_msg <- paste("Improve this text. Focus on", style, ". Show improved version.")
      cat(ask_llm(text, sys_msg, temperature = 0.3))
    },
    "4" = {
      question <- readline("Your question: ")
      cat(ask_llm(question))
    },
    cat("Invalid option. Try again.\n")
  )
}
```

---
class: exercise_slide

## `r ICONS$desktop` Exercise: Add One Menu Item

.font90[
Extend the study assistant with:

`5 - Translate text`

Use the `translate_text()` function from earlier. Keep the menu logic consistent with the existing options.
]

<br>
.content-box-white[
This is a realistic extension: no new API concept, but more control flow.
]

???
Joint solution:

```r
show_menu <- function() {
  cat("\n=== Study Assistant ===\n")
  cat("1 - Summarize text\n")
  cat("2 - Generate flashcards\n")
  cat("3 - Improve my writing\n")
  cat("4 - Ask a question\n")
  cat("5 - Translate text\n")
  cat("q - Quit\n")
  readline("Choose an option: ")
}
```

Add this branch to `switch()`:

```r
"5" = {
  text <- readline("Enter text to translate: ")
  lang <- readline("Target language (default: German): ")
  if (lang == "") lang <- "German"
  cat(translate_text(text, target_language = lang))
}
```

---
## Error Handling for API Calls

.font90[
API calls can fail because of:

- missing or invalid keys
- rate limits
- network problems
- unavailable models
- malformed inputs
]

<br>
.content-box-white[
Production-facing helper functions should catch errors and return useful messages.
]

???
Live code:

```r
ask_llm_safe <- function(user_message,
                         system_message = "You are a helpful assistant.") {
  tryCatch(
    {
      chat <- chat_groq(system_prompt = system_message, echo = "none")
      chat$chat(user_message)
    },
    error = function(e) {
      msg <- conditionMessage(e)
      if (grepl("401|authentication|unauthorized", msg, ignore.case = TRUE)) {
        "Error: Invalid API key. Check your .Renviron file."
      } else if (grepl("429|rate.limit", msg, ignore.case = TRUE)) {
        "Error: Too many requests. Wait a moment and try again."
      } else if (grepl("connection|network|timeout", msg, ignore.case = TRUE)) {
        "Error: Cannot reach the API. Check your internet connection."
      } else {
        paste("Error:", msg)
      }
    }
  )
}
```

---
## Best Practices

.pull-left[
**Prompting**

- state the role
- specify audience
- specify output format
- include constraints
- test edge cases
]

.pull-right[
**Engineering**

- keep keys out of code
- capture output for scripts
- monitor `token_usage()`
- use low temperature for extraction
- review model outputs
]

---
## What We Learned

.font90[
.pull-left[
**Concepts**

- APIs as remote function calls
- chat objects and history
- system prompts
- parameters and providers
- multimodal input
]

.pull-right[
**Tools**

- simple Groq-based LLM requests
- prompt experiment helper function
- LLM summarizer
- LLM flashcard generator
- LLM translator
- structured extractor
- study assistant skeleton
]
]

<div style="clear:both"></div>

.font90[
Further reading: [ellmer documentation](https://ellmer.tidyverse.org), [prompt design](https://ellmer.tidyverse.org/articles/prompt-design.html), [structured data](https://ellmer.tidyverse.org/articles/structured-data.html)
]

