---
title: "Exercise Solutions data wrangling"
author: 'Jens Klenke' 
output: 
  html_document:
   css: ["../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css 
# https://github.com/gadenbuie/xaringanExtra
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"),
  rmarkdown::html_dependency_font_awesome() 
  )
```

```{r packages}
library(tidyverse)
library(nycflights13)
library(ggplot2)
library(readr)
library(here)
```


1.  Load the <code>nycflights13::flights</code> data set. Does it come formatted as a tibble? 

```{r}
data("flights")
class(flights)
```

2. Find all flights that 

  - Had an arrival delay of two or more hours.

```{r}
filter(flights, arr_delay >= 120)
```

  - Flew to Houston (IAH or HOU)


```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

  - Were delayed by at least an hour, but made up over 30 minutes in flight.

```{r}
filter(flights, dep_delay >= 60 & arr_delay <= dep_delay - 30)
```

  - Departed between midnight and 6am.

```{r}
filter(flights , between(dep_time, 0, 600) | dep_time == 2400)
```
  
3. Which flights took the longest and which the shortest?

```{r}
slice(flights, which.max(air_time))
slice_max(flights, air_time)
```

```{r}
slice(flights, which.min(air_time))
dplyr::slice_min(flights, air_time)
```


4.  Brainstorm as many ways as possible to select <code>dep_time</code>, <code>dep_delay</code>, <code>arr_time</code>, and <code>arr_delay</code>.

```{r}
selection <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
flights[ ,selection]

flights %>% select(selection)
```

5. Create a new variable <code>first_flight</code> which takes the value <code>TRUE</code> if it is the first flight of the day. 

```{r}
flights <-  mutate(flights, first_flight = if_else(day - lag(day) != 0 | is.na(day - lag(day)), TRUE, FALSE )) 
filter(flights, first_flight == TRUE)

flights %>%
  group_by(year, month, day) %>%
  mutate(first_flight = if_else(row_number() == 1, TRUE, FALSE)) %>%
  select(first_flight)
```

6. Find out in which district the most incidents were reported.

```{r}
crime <- readr::read_csv(here::here('SoSe_2022/datasets/BostonCrime.csv'))

crime %>% 
  group_by(DISTRICT) %>%
  summarise(N = n()) %>% 
  arrange(desc(N))
```
7. Create an ordered list of the top ten districts according to the number of drug violations. Start with the district with most drug violations. 

```{r}
crime %>% 
  filter(OFFENSE_CODE_GROUP == "Drug Violation") %>% 
  group_by(DISTRICT) %>% 
  summarise(N = n()) %>% 
  arrange(desc(N)) 
```

8. What was the most dangerous hour of the week according to the number of homicides (<code>OFFENSE_CODE_GROUP == 'Homicide'</code>)? 

```{r}
crime %>% 
  filter(OFFENSE_CODE_GROUP == "Homicide") %>% 
  group_by(HOUR, DAY_OF_WEEK) %>%
  summarise(N = n()) %>% 
  arrange(desc(N))
```

9. Calculate for each district the proportion of motor vehicle accidents to all reported incidents (<code>OFFENSE_CODE_GROUP == 'Motor Vehicle Accident Response'</code>). 

```{r}
 crime %>% 
  group_by(DISTRICT) %>%  
  transmute(prop_accidents = sum(OFFENSE_CODE_GROUP == "Motor Vehicle Accident Response") / n()) 
```

10. Sort the data set chronologically by year and month. Then take the first 10.000 observations and compute the proportion of incidents where a shooting was involved for each district. Now take a random sample of size 10.000. Is there a noteworthy difference? <i>Hint:</i> There is a variable called <code>SHOOTING</code>.

```{r}
crime  %>%
  arrange(desc(YEAR), desc(MONTH)) %>%
  slice(1:10000) %>%
  mutate(SHOOTING = ifelse(is.na(SHOOTING), 0, SHOOTING)) %>%
  summarise(porpotion = mean(SHOOTING == "Y"))
```


```{r}
set.seed(1245)

crime  %>%
  slice_sample(n = 10000) %>%
  mutate(SHOOTING = ifelse(is.na(SHOOTING), 0, SHOOTING)) %>%
  summarise(porpotion = mean(SHOOTING == "Y"))
```

11. Compute the TB rate for table2, table4a and table4b.

```{r}
table2[table2$type == "cases", "count"] / table2[table2$type == "population", "count"]
```

```{r}
table4a[ ,-1] / table4b[ ,-1]
```

12. Recreate the plot given by the code below using `table2` instead of `table1`. 

```{r}
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```




```{r}
table2 %>% 
  filter(type == "cases") %>%
  ggplot( aes(x = year, y = count)) +
    geom_line(aes( group = country)) +
    geom_point(aes(color = country))
```
