---
title: "Exercise Solutions data wrangling"
author: 'Jens Klenke & Martin Schmelzer'
output: 
  html_document:
   css: ["../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
   number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = F)
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
source(here::here("SoSe_2026/utils.R"))
fontawesome::fa_html_dependency()
```

```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

```{r packages}
library(tidyverse)
library(nycflights23)
library(ggplot2)
library(readr)
library(here)
```

## Load the <code>nycflights23::flights</code> data set. Does it come formatted as a tibble? 

```{r}
data("flights")
class(flights)
nrow(flights)
```

## Find all flights that 

- **Had an arrival delay of two or more hours.**

```{r}
filter(flights, arr_delay >= 120)
```

- **Flew to Houston (IAH or HOU)**


```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

- **Were delayed by at least an hour, but made up over 30 minutes in flight.**

```{r}
filter(flights, dep_delay >= 60 & arr_delay <= dep_delay - 30)
```

- **Departed between midnight and 6am.**

```{r}
filter(flights, between(dep_time, 0, 600) | dep_time == 2400)
filter(flights, when_any(
  between(dep_time, 0, 600),
  dep_time == 2400
))
```
  
## Which flights took the longest and which the shortest?

```{r}
flights %>%
  arrange(desc(air_time)) %>%
  slice(1)
slice(flights, which.max(air_time))
slice_max(flights, air_time)
```

```{r}
slice(flights, which.min(air_time))
dplyr::slice_min(flights, air_time)
```


## Brainstorm as many ways as possible to select <code>dep_time</code>, <code>dep_delay</code>, <code>arr_time</code>, and <code>arr_delay</code>.

```{r}
# Basic selection
select(flights, dep_time, dep_delay, arr_time, arr_delay)

# BY character vector
selection <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
flights[ ,selection]
select(flights, selection)

# String matching
select(flights, starts_with("dep"), starts_with("arr"))
select(flights, matches("^(dep|arr)_(time|delay)$"))
```

## Create a new variable <code>first_flight</code> which takes the value <code>TRUE</code> if it is the first flight of the day. 

```{r}
flights %>%
  mutate(first_flight = if_else(day - lag(day) != 0 | is.na(day - lag(day)), TRUE, FALSE)) %>%
  filter(first_flight == TRUE)

flights %>%
  group_by(year, month, day) %>%
  mutate(first_flight = if_else(row_number() == 1, TRUE, FALSE)) %>%
  filter(first_flight)
```

## Create a new variable called <code>speed</code> (in miles per hour) using <code>distance</code> and <code>air_time</code>. Note that <code>air_time</code> is in minutes. Identify the 10 fastest flights in the dataset and check if they belong to a specific carrier.


```{r}
flights %>%
  mutate(speed = distance / (air_time / 60)) %>%
  slice_max(speed, n = 10) %>%
  select(carrier, origin, dest, speed) %>%
  count(carrier)
```




## Find out in which district the most incidents were reported.

```{r}
crime <- readr::read_csv(here::here('SoSe_2026/datasets/BostonCrime.csv'))

crime %>% 
  group_by(DISTRICT) %>%
  summarise(N = n()) %>% 
  arrange(desc(N))
```

## Create an ordered list of the top ten districts according to the number of drug violations. Start with the district with most drug violations. 

```{r}
crime %>% 
  filter(OFFENSE_CODE_GROUP == "Drug Violation") %>% 
  group_by(DISTRICT) %>% 
  summarise(N = n()) %>% 
  arrange(desc(N)) 
```

## What was the most dangerous hour of the week according to the number of homicides (<code>OFFENSE_CODE_GROUP == 'Homicide'</code>)? 

```{r}
crime %>% 
  filter(OFFENSE_CODE_GROUP == "Homicide") %>% 
  group_by(HOUR, DAY_OF_WEEK) %>%
  summarise(N = n()) %>% 
  arrange(desc(N))
```

## Calculate for each district the proportion of motor vehicle accidents to all reported incidents (<code>OFFENSE_CODE_GROUP == 'Motor Vehicle Accident Response'</code>). 

```{r}
crime %>% 
  group_by(DISTRICT) %>%  
  transmute(prop_accidents = sum(OFFENSE_CODE_GROUP == "Motor Vehicle Accident Response") / n()) 
```

## Sort the data set chronologically by year and month. Then take the first 10.000 observations and compute the proportion of incidents where a shooting was involved for each district. Now take a random sample of size 10.000. Is there a noteworthy difference? <i>Hint:</i> There is a variable called <code>SHOOTING</code>.

```{r}
crime %>%
  arrange(desc(YEAR), desc(MONTH)) %>%
  slice(1:10000) %>%
  mutate(SHOOTING = replace_na(SHOOTING, "0")) %>%
  summarise(porpotion = mean(SHOOTING == "Y"))
```


```{r}
set.seed(1245)

crime  %>%
  slice_sample(n = 10000) %>%
  mutate(SHOOTING = replace_na(SHOOTING, "0")) %>%
  summarise(propotion = mean(SHOOTING == "Y"))
```

Selection bias -> temporal bias -> random sampling is crucial.
Recent value relevant for forecasting -> population sample relevant for understanding base line

## Compute the TB rate for table2 and table4a + table4b.

```{r}
table2 %>%
  tidyr::pivot_wider(names_from = type, values_from = count) %>%
  dplyr::mutate(rate = cases / population)
```

```{r}
table4a %>%
  tidyr::pivot_longer(cols = c(`1999`, `2000`), 
                      names_to = "year", 
                      values_to = "cases") %>%
  dplyr::left_join(table4b %>%  
                      tidyr::pivot_longer(cols = c(`1999`, `2000`), 
                                          names_to = "year", 
                                          values_to = "population"), 
                   by = c("country", "year")) %>%
  dplyr::mutate(rate = cases / population)
```




## Take the <code>tidyr::who</code> and <code>tidyr::who2</code> dataset.

- **The who2 dataset is missing <code>iso2</code> and <code>iso3</code> codes. Create a lookup table using the original <code>tidyr::who</code> dataset that contains only the unique combinations of <code>country</code>, <code>iso2</code>, and <code>iso3</code>.**

- **Use a left join to add these codes to your who2 dataset.**

- **Once joined, pivot the dataset so that all the tuberculosis measurement columns (from <code>sp_m_014</code> to <code>rel_f_65</code>) are moved into two columns: <code>category_code</code> and <code>count</code>.**

- **Use <code>separate()</code> to split the <code>category_code</code> into three distinct variables: <code>method</code>, <code>sex</code>, and <code>age_group</code>.**

- **The age_group codes are a bit messy (e.g., 014, 1524, 65). Use <code>mutate()</code> and <code>case_when()</code> to rename 014 to 00-14, 65 to 65+ and make all other groups use a hyphen (-) to seperate their limits.**

```{r}
iso_lookup <- who %>%
  distinct(country, iso2, iso3)

who2_enriched <- who2 %>%
  left_join(iso_lookup, by = "country") 

who2_enriched_sep <- who2_enriched %>%
  pivot_longer(sp_m_014:rel_f_65, names_to = "category_code", values_to = "count") %>%
  separate(category_code, into = c("method", "sex", "age"), sep = "_")

who2_enriched_sep %>%
  mutate(age = case_when(
    age == "014" ~ "00-14",
    age == "1524" ~ "15-24",
    age == "2534" ~ "25-34",
    age == "3544" ~ "35-44",
    age == "4554" ~ "45-54",
    age == "5564" ~ "55-64",
    age == "65" ~ "65+",
    TRUE ~ age
  ))
  
# OR
who2_enriched_sep %>%
  mutate(age = case_when(
    age == "014" ~ "00-14",
    age == "65"  ~ "65+",
    str_length(age) == 4 ~ str_replace(age, "(\\d{2})(\\d{2})", "\\1-\\2"),
    TRUE ~ age
  ))

```



<!-- ## Recreate the plot given by the code below using `table2` instead of `table1`. 

```{r}
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```


```{r}
table2 %>% 
  filter(type == "cases") %>%
  ggplot(aes(x = year, y = count)) +
    geom_line(aes( group = country)) +
    geom_point(aes(color = country))
```
-->