---
title: "Advanced R for Econometricians"
subtitle: "Data Wrangling"
author: 
  - "Martin Arnold"
  - "Martin Schmelzer"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/navigate_home.js", "../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
class: title-slide title-wrangling center middle
```{r setup, echo = F, eval = T, warning = F}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE)
source(here::here("SoSe_2026/utils.R"))
fontawesome::fa_html_dependency()
```

```{r moon_reader, include=FALSE, eval = FALSE}
# setup for moon reader
# copy it to console
options(servr.interval = 0.5)
xaringan::inf_mr()
```



# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`

---
class: top, left
## Tibble

.font90[
Tibbles are the main data structure used in the `tidyverse` and can be viewed as a modern version of the data frame. 

- `tibble()` creates a new tibble from scratch.
]


.blockquote.exercise.font90[
```{r, echo = TRUE, eval = TRUE, warning=F, message=F}
library(tidyverse)
new_tibble <- tibble(x = 1:3, y = letters[1:3])
```
]

--

.font90[
- `as_tibble()` creates a tibble from an existing object (e.g. a data frame or a matrix).
]

.blockquote.exercise.font90[
```{r, echo = TRUE, eval = TRUE}
    a_data.frame <- data.frame(x = 1:3, y = letters[1:3])
    a_tibble     <- as_tibble(a_data.frame)
```
]

--


.font90[
- Everything that works for data frames also works with tibbles because
]
.blockquote.exercise.font90[
```{r, echo = TRUE, eval = TRUE}
class(a_tibble)
```
]

---
class: top, left

## Tibble vs. data.frame
- Why should we use tibbles instead of data frames? See the following comparison.  

.blockquote.exercise.font90[
```{r, echo = TRUE}
data(mtcars) 
mtc_tib <- as_tibble(mtcars)
```
]

--

- different print method

.blockquote.exercise.font90[
```{r, echo = TRUE, eval = FALSE}
print(mtcars)
print(mtc_tib)
```
]

???
- Shows data types
- Shows dimensions

--

- more consistent subsetting behavior

???
mtc_tib[, 1] stays a tibble

.blockquote.exercise.font80[
#### `r ICONS$desktop` Example

```{r, echo = TRUE, eval = FALSE}
mtcars[, 1:2] 
mtcars[, 1]    
mtc_tib[, 1:2]
mtc_tib[, 1] 
```
]
---
## Tibble vs. data.frame


- lazy and sequential evaluation 

.blockquote.exercise[
#### `r ICONS$desktop` Example
```{r, echo = TRUE, eval = FALSE}
data.frame(x = 1:3, y = x^2)  # not working
tibble(x = 1:3, y = x^2)
```
]


--

- higher flexibility with nested data

.blockquote.exercise[
#### `r ICONS$desktop` Example
```{r, echo = TRUE, eval = FALSE}
data_df  <- data.frame(days = 1:3,
                       data = list(mtcars)) # not working

data_tib <- tibble(days = 1:3, 
                   data = list(mtcars))
```
]

- For more see `vignette("tibble")`.


---
class: top, left
background-image: url(../assets/dplyr.svg)
background-size: 150px
background-position: 97.5% 8%

## dplyr 

<br>

The `dplyr` package is the most important package of the `tidyverse` when it comes to data manipulation and transformation. 

--

The core characteristics of the package:

- focus on data frames (data frame in &rarr; data frame out)
- verbose compared to base R which leads to longer but easier to read code
- consistent syntax. 

???
With every update, the developers try to increase the readability and decrease code length

--

You can compare the functionality to that of a  relational database which provides methods to work with single tables and multiple tables. 

---
class: top, left
## Single Table Functions

First, we look at functions that work on a single table. 
 
- `filter()`/`filter_out()` to select/deselect rows based on their values.

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, hp >= 100 & mpg >= 20)
filter_out(mtc_tib, hp >= 100 & mpg >= 20)
```
]

???
filter_out was added in the last major update
filter_out(mtc_tib, hp < 100 | mpg < 20)

--

- `slice()` to select rows by position. 

.blockquote.exercise[
```{r, eval = FALSE, echo = TRUE}
slice(mtc_tib, 10:15)
```
]

--

- `select()`/`rename()` to select/rename variables based on their names.

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
select(mtc_tib, cyl, hp)
```
]

---
class: top, left
## Single Table Functions

- `arrange()` to reorder rows.

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
arrange(mtc_tib, cyl)
arrange(mtc_tib, desc(cyl))
```
]

--

- `relocate()` to reorder columns (variables).

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
relocate(mtc_tib, gear)
relocate(mtc_tib, gear, .after = last_col())
relocate(mtc_tib, gear, .before = cyl)
```
]


---
class: top, left
## Single Table Functions

- `mutate()` and `transmute()` to add new variables that are functions of existing variables. 

.blockquote.exercise[

```{r, echo = TRUE, eval = FALSE}
mutate(mtc_tib, gpm = 1/mpg)
transmute(mtc_tib, gpm = 1/mpg) 
```
]

--

- `summarise()` to condense multiple values to a single value.

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
summarise(mtc_tib, mean_hp = mean(hp), max_mpg = max(mpg))
```
]

---
class: top, left
## Single Table Functions

- `slice_sample()` to take random samples

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
slice_sample(mtc_tib, n = 10)
slice_sample(mtc_tib, prop = 0.1)
```
]

--

- `distinct()` removes duplicate rows

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
tibble_with_duplicates    <- tibble(x = c(1, 1, 2, 3), 
                                    y = c(4, 4, 5, 6))
tibble_without_duplicates <- distinct(tibble_with_duplicates)
```
]

--

- `count()` and `tally()` count rows.  

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
count(mtc_tib); count(mtc_tib, cyl)
tally(mtc_tib); tally(mtc_tib, cyl)
```
]

---
class: top, left

.font80[
## Filter 

The function `filter()` selects rows based on their values. 

- Use comparison operators `==`, `>`, `>=`, `<`, `<=`, and `!=` to filter rows based on a logical condition. 

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, hp == 100)
```
]]

???
Lets taker a closer look 

--

.font80[
- Use logical operators such as  `&` and `|` to combine multiple conditions. 

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, hp > 100 & cyl == 4)
filter(mtc_tib, hp > 100 | cyl == 4)
```
]]

--

.font80[
- Additional arguments to `filter()` are combined in the back end with `&` which is the same as the above. 

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, hp > 100, cyl == 4)
```
]]

--

.font80[
- For the common operation `x >= left & x <= right` there is the shorthand `between(x, left, right)`.

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, x >= 90 & x <= 110)
filter(mtc_tib, between(hp, 90, 110))
```
]]

---
class: top, left

.font80[
## Filter 

- Since `dplyr 1.2.0` you easily invert your filter by replacing `filter()` with `filter_out()`

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, hp > 110 | cyl == 4)
filter_out(mtc_tib, hp > 110 | cyl == 4)
```
]

- Also new are the helper functions `when_any` and `when_all`

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
filter(mtc_tib, when_any(
  mpg > 25 & wt < 2.5,
  hp > 200 & wt > 3.5
))

filter(mtc_tib, when_all(
  hp > 150,
  mpg > 15,
  cyl == 8 & am == 0
))
```
]]


???
You might wonder if you need them, you could just use & and |, but the main argument here is readability.
If you have a big data.frame and your business logic requires many different logical comparisons, this might come in handy.


---
## Select

.font90[
The function `select()` selects columns based on their name. 

- The function can handle standard and non-standard evaluation (more about this later). 

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
select(mtc_tib, disp, vs) 
select(mtc_tib, "disp", "vs") 
```
]]

--

.font90[
- You can also work with character vectors. 

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
select(mtc_tib, c("disp", "vs"))
```
]]

--

.font90[
- Use `-` to select all except the provided columns.

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
select(mtc_tib, -disp, -vs) 
select(mtc_tib, -c("disp", "vs")) 
```
]

- There is a whole bunch of helper functions for `select()` on the next slides. 
]

---
class: top, left

## Helper Functions for `select()`

.font90[
- Select columns whose name contains a character string.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, contains("."))
```
]]

--

.font90[
- Select columns whose name ends with a character string.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, ends_with("Length"))
```
]]

--

.font90[
- Select columns whose name matches a regular expression.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, matches(".t."))
```
]]

--

.font90[
- Select columns named `x1`, `x2`, `x3`, `x4`, `x5`.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, num_range("x", 1:5))
```
]]

---
class: top, left
## Helper Functions for `select()`

.font90[
- Select all columns from `Sepal.Width` to `Petal.Width`.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, Sepal.Width:Petal.Width)
```
]]

--

.font90[
- Select columns whose names are in a group of names.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, one_of(c("Species", "Genus")))
```
]]

--

.font90[
- Select columns whose name starts with a character string.

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
select(iris, starts_with("Sepal"))
```
]]
---
## Mutate

.font80[
The function `mutate()` adds new columns or replaces old columns based on existing columns. 

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
mutate(mtc_tib,
       gpm = 1/mpg,
       disp = disp - mean(disp)) # replaces disp 
```
]]

--

.font80[
- Use `lag()` and `lead()` to access values that are not in the same row (e.g. for time series data).

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
rw <- tibble(t = 1:10, y = cumsum(rnorm(10)))
mutate(rw, increments =  y - lag(y, k = 1))
```
]]

--

.font80[
- Use `min_rank()` to add a column with the rank (`?min_rank` gives you an overview of further ranking functions).

.blockquote.exercise[
```{r, echo=TRUE, eval = FALSE}
mutate(mtc_tib,
       hp_rank_smallest_first = min_rank(hp),
       hp_rank_largest_first  = min_rank(desc(hp)))
```
]]

---
## Summarise

.font80[

`summarise()` can be used to collapse columns to a single row used with e.g.: 
- base R functions such as `mean()`, `sd()`, `max()`
- dplyr specific functions for
  - position `first()`, `nth()`, `last()`
  - counts `n()`, `n_distinct()`

]

???
This is where it gets interesting, usuallly the last coommands

--

.blockquote.exercise[.font80[

```{r, echo = TRUE, eval = FALSE}
summarise(mtc_tib,
          mean_hp    = mean(hp),
          max_wt     = max(wt), 
          n          = n(),            
          third_cyl  = nth(cyl, 3))
```
]]

???
Lets look at that

--

.font80[

- functions you have written yourself 

.blockquote.exercise[
```{r, echo = TRUE, eval = FALSE}
my_variance <- function(x) 1/(n()-1) * sum((x - mean(x))^2)
summarise(by_cyl, my_variance(mpg))
```

]]

???
my_variance(1:10)
n() works only inside data-masking verbs


---
## Summarise

- You cannot use `summarise()` with functions that return more than one value per group. 

.blockquote.exercise[
```{r, eval = FALSE, echo = TRUE}
two_return_values <- function(x) c(mean(x), sd(x))
summarise(by_cyl, two_return_values(mpg))
```
]

--

- The function can take multiple variables or fixed inputs.

.blockquote.exercise[
    ```{r, echo = TRUE, eval = FALSE}
two_input_values <- function(x, y) sum(x + y)
summarise(by_cyl, two_input_values(hp, qsec)) # variable inputs
summarise(by_cyl, two_input_values(hp, 1))    # fixed input
    ```
]

---
class: exercise_slide
## Exercises

<ol start="1"> 
<li>  Load the <code>nycflights13::flights</code> data set. Does it come formatted as a tibble? </li>
<li>  Find all flights that </li>
  <ul>
      <li> Had an arrival delay of two or more hours.</li>
      <li> Flew to Houston (IAH or HOU).</li>
      <li> Were delayed by at least an hour, but made up over 30 minutes in flight.</li>
      <li> Departed between midnight and 6am.</li>
  </ul>
<li> Which flights took the longest and which the shortest? </li>
<li> Brainstorm as many ways as possible to select <code>dep_time</code>, <code>dep_delay</code>, <code>arr_time</code>, and <code>arr_delay</code>. </li>
<li>  Create a new variable <code>first_flight</code> which takes the value <code>TRUE</code> if it is the first flight of the day. </li>
<li> Create a new variable called speed (in miles per hour) using <code>distance</code> and <code>air_time</code>. Note that <code>air_time</code> is in minutes. Identify the 10 fastest flights in the dataset and check if they belong to a specific carrier.</li>
</ol>

```{r, echo=FALSE, eval = FALSE}
library(tidyverse)
library(nycflights13)
filter(flights, day == 1)
```



---
class: top, left
## Grouping

- To exploit the full potential of `summarise()` we pair it with `group_by()`. 

--

.blockquote.exercise[
```{r, echo = TRUE, results='markup'}
by_cyl <- group_by(mtc_tib, cyl)
summarise(by_cyl, mean(mpg))
```
]

--

- `.by` as an alternative to `group_by`:

.blockquote.exercise[
```{r, echo = TRUE, results='markup'}
summarise(mtc_tib, mean(mpg), .by = cyl)
```
]

--

- Use `ungroup()` to remove the grouping, when you want to work on the whole data set again. 




---
class: top, left
## Grouping

It is also possible to use `group_by()` together with a lot of other single table operations.

--
 
.blockquote.exercise[
#### `r ICONS$desktop` Example

```{r, echo = TRUE, eval = FALSE}
by_cyl <- group_by(mtc_tib, cyl)

# group mean is used for centering
mutate(by_cyl, centered_mpg = mpg - mean(mpg))   
transmute(by_cyl, centered_mpg = mpg - mean(mpg))

# return only first 3 rows for each group
slice(by_cyl, 1:3)  

# counts the number of rows for each group
count(by_cyl)   

# returns all groups with more than 10 rows
filter(by_cyl, n() > 10)  

# sorts one group after the other
arrange(by_cyl, disp, .by_group = TRUE)                        
```
]

---
class: top, left
## The Pipe

Usually, we combine multiple operations, e.g. first we filter, then we mutate, group and finally summarise. 

--

We could do it like this: 

.blockquote.exercise[ 

#### `r ICONS$incorrect` Bad
```{r, echo = TRUE}
filtered   <- filter(mtc_tib, cyl <= 8)
mutated    <- mutate(filtered, gpm = 1/mpg)
grouped    <- group_by(mutated, gear)
summarized <- summarise(grouped, count = n(), 
                        mean_hp = mean(hp),
                        mean_gpm = mean(gpm)) 
```
]

--

<br>

This code has the disadvantage that we need to save intermediate results that we do not care about. 

---
class: top, left
## The Pipe

We could nest all those functions to get rid of the intermediate results.

.blockquote.exercise[ 
#### `r ICONS$desktop` Example
```{r, echo = TRUE}
summarise(group_by(mutate(filter(mtc_tib, cyl <= 8), gpm = 1/mpg)), 
          count = n(), 
          mean_hp = mean(hp),
          mean_gpm = mean(gpm))
```
]

--

<br>
However, readability gets worse and debugging will become much harder. 
---
class: top, left
## The Pipe

.font90[
That is where the pipe operator `%>%` from the `magrittr` package comes in handy.
The value on the left side of the operator becomes the input to the function on the right side. ]

--

.blockquote.exercise.font80[ 
#### `r ICONS$desktop` Example
```{r, echo = TRUE, eval = FALSE}
mtc_tib %>% 
  mutate(gpm = 1/mpg) %>%
  summarise(count = n(), 
            mean_hp = mean(hp),
            mean_gpm = mean(gpm), 
            .by = gear) 
```
]

???
and made use of .by

--

.font90[
The pipe operator is not limited to `dplyr` or the `tidyverse`

.blockquote.exercise[ 
```{r, echo = TRUE, eval = FALSE}
1:6 %>% mean
```
]]

--

.font90[
The package `magrittr` offers different pipe operators for more flexbility:

- `%<>%` combines the pipe and `<-`
- `%T>%` to do something on the side, i.g. generating a plot
- `%$%` collapse the `data.frame` class
]

???
Originally the pipe operators were developed in magrittr

Examples:

#### Overwriting an object
mtc_tib %<>% dplyr::summarise(mean(hp))



#### by side plotting
cars_1 <- cars %>%
  dplyr::select(hp, mpg) %T>%
  plot %>%
  dplyr::summarise(mean(hp))


#### select from data.frame
cars %>%
  dplyr::select(hp, mpg) %$%
  cor(hp, mpg)

  


```{r, eval = FALSE, cache=F}
library(data.table)
library(tidyverse)
crime <- fread("https://data.boston.gov/dataset/6220d948-eae2-4e4b-8723-2dc8e67722a3/resource/12cb3883-56f5-47de-afa5-3b1cf61b257b/download/tmppj4rb047.csv")
crime %>% group_by(DISTRICT) %>% summarise(N = n()) %>% arrange(desc(N)) 
crime %>% dplyr::filter( OFFENSE_CODE_GROUP == "Drug Violation") %>% group_by(OFFENSE_CODE_GROUP, DISTRICT) %>% summarise(N = n()) %>% arrange(desc(N)) 


d1  <- crime %>% arrange(YEAR) %>% slice(1:10000) %>% group_by(DISTRICT, SHOOTING) %>% summarise(N = n()) 
d2  <- crime %>% arrange(YEAR) %>% sample_n(10000) %>% group_by(DISTRICT) %>% summarise(N = n()) 
plot(x = factor(d1$DISTRICT), d1$N)
points(x = factor(d2$DISTRICT), d2$N)
```


```{r, eval = FALSE}
plot(table(crime$DISTRICT[1:10000])/sum(table(crime$DISTRICT[1:10000])))
points(table(crime$DISTRICT)/sum(table(crime$DISTRICT)), col = alpha("red", 0.4), lty = 3, lwd = 5)
```


---
class: top, left
## Scoped Functions 

We often perform operations on all or several columns at once. Most helpful are the selection auxiliary functions of the `dplyr` package.

- Use `across()` to determine the variable(s) and summarise function(s)  on specific columns.

.blockquote.exercise[ 
```{r, eval = FALSE, echo = TRUE}
iris %>% summarise(across(c(Sepal.Length, Sepal.Width), mean))
iris %>% summarise(across(c(Sepal.Length, Sepal.Width), ~ mean(.x, na.rm = TRUE)))
# summarise with multiple functions 
iris %>% summarise(across(c(Sepal.Length, Sepal.Width), list(mean = mean, sd = sd)))
```
]

--

- Inside the `across()` function we can use the helper functions for select which we already saw on the previous slides. 

.blockquote.exercise[ 
```{r, eval = FALSE, echo = TRUE}
iris %>% summarise(across(ends_with('Width'), mean))
```
]

???

The ~ operator in R creates formula. So ~ lm(mpg ~ wt, data = .) is a formula. Formulas are useful in R because they prevent immediate evaluation of symbols. For example you can define


purrr-style formula? 


---
class: top, left
## Advanced Topics 

.font90[
If `summarise()` is too limited, because it can only handle functions that have a single 
return value, use 
  - `group_modify(.tbl, .f, ...)` to return a grouped tibble. In that case `.f` must return a data frame.
  - `group_map(.tbl, .f, ...)` to return a list of results from calling `.f` on each group.
  - `group_walk(.tbl, .f, ...)` to call `.f` for side effects and returns the input `.tbl`, invisibly.

]

--

.font90[
Those functions are closely related to the functions in the `purrr` package which we 
will talk about later. Here we only give a short example of how they can be used. 

.code80[.blockquote.exercise[ 
```{r, echo = TRUE, eval = FALSE}
mtcars %>%
  group_by(cyl) %>% 
  group_map(~ lm(mpg  ~ hp, data = . ))

mtcars %>%
  group_by(cyl) %>% 
  group_modify(~ broom::tidy(lm(mpg  ~ hp, data = . )))

par(mfrow = c(2,2))
mtcars %>%
  group_by(cyl) %>% 
  group_walk(~ plot(mpg  ~ hp, data = . ))
```
]]]

---
class: exercise_slide

## Exercises

Use `dplyr` on the Boston crime data to:

<ol start="6"> 
<li>  Find out in which district the most incidents were reported.  </li>
<li>  Create an ordered list of the top ten districts according to the number of drug violations. Start with the district with most drug violations. </li>
<li>  What was the most dangerous hour of the week according to the number of homicides (<code>OFFENSE_CODE_GROUP == 'Homicide'</code>)?  </li>
<li>  Calculate for each district the proportion of motor vehicle accidents to all reported incidents (<code>OFFENSE_CODE_GROUP == 'Motor Vehicle Accident Response'</code>).   </li>
<li>  Sort the data set chronologically by year and month. Then take the first 10.000 observations and compute the proportion of incidents where a 
shooting was involved for each district. Now take a random sample of size 10.000. Is there a noteworthy difference? <i>Hint:</i> There is a variable called <code>SHOOTING</code>.
  </li>
</ol>


???


dataset:
https://www.kaggle.com/AnalyzeBoston/crimes-in-boston?select=crime.csv

---
class: top, left
## Functions for 2 Tables

- Two data frames

.blockquote.exercise[ 
```{r, echo = TRUE, eval=T, results='markup'}
band_members   
band_instruments
```
]

---
class: top, left
## Left Join

A left join keeps all observations from the left table, and matches records from the right table based on one (or multiple) columns (`by`).
.blockquote.exercise[ 
```{r, echo = TRUE, results='markup'}
band_members %>% left_join(band_instruments, by = "name")
```
]

--

- Sometimes you may wish to keep all entrances of the right `data.fame`. You can do so with `right_join()`

---
class: top, left
## Inner Join

An inner join matches rows from the left and the right table but only keeps the rows where both tables have matching rows in the `by` column.
.blockquote.exercise[ 
```{r, echo = TRUE, results='markup'}
band_members %>% inner_join(band_instruments, by = "name")
```
 ]
---
class: top, left
## Full Join

A full (outer) join matches rows from the left and the right table and keeps all observation

.blockquote.exercise[ 
```{r, echo = TRUE, results='markup'}
band_members %>% full_join(band_instruments, by = "name")
```
]
---
class: top, left
## Semi Join

A semi join keeps all observations of the left table where there are matching values in the right table.
It keeps only columns from the left. 

.blockquote.exercise[ 
```{r, echo = TRUE, results='markup'}
band_members %>% semi_join(band_instruments, by = "name")
```
]

---
class: top, left
## Anti Join
An anti join returns all rows from the first table where no matches are found in the second table.

.blockquote.exercise[  
```{r, echo = TRUE, results='markup'}
band_members %>% anti_join(band_instruments, by = 'name')
```
]

---
class: top, left
## Reshaping Data with tidy R

The goal of `tidyr` is to help you create tidy data. Tidy data is data where:

- each variable is a column
- each observation is a row
- each value is a cell.

We will talk about 

- `pivot_longer()` takes multiple columns, and gathers them into _key-value_ pairs: it makes “wide” data longer.
- `pivot_wider()` takes multiple _value_ columns and one _key_ column, and spreads them across multiple columns: it makes “long” data wider.
- `separate()` takes a single column and turns it into multiple columns based on a separator.  
- `unite()` takes multiple columns and pastes them together into one column.  

---
class: top, left
## Tidy Data

.blockquote.exercise[
```{r, echo = TRUE}
tidyr::table1
```
]

--

Each observation is in one row and each variable is one column. 

---
class: top, left
## Untidy data

.blockquote.exercise[
```{r, echo = TRUE}
tidyr::table2
```
]

--

Observations are scattered across multiple rows

???
The data.frame is in a long format.

---
class: top, left
## Spreading (Widening)

The function `tidyr::pivot_wider()` main arguments are

- `names_from`, containing the variable names
- `values_from`, containing the values for the variables in the `names_from` column. 

.blockquote.exercise[
```{r, echo = TRUE}
table2 %>%
    pivot_wider(names_from = type, values_from = count)
```
]

--

See the help file for more information

???
Many more arguments that allow to customize column naming or how NULL values should be handled

---
## Untidy Data

.blockquote.exercise[
```{r, echo = TRUE}
tidyr::table4b
```
]

--

- Variables year and population are spread across multiple columns and column names are values. 

---
## Gathering (Lengthening)

.font90[
The function `pivot_longer()` main arguments are

- `cols` the column names which are actual values
- `names_to`, the name of the newly created variable resulting from the column name
- `values_to`, the name of the newly created variable.
]

--

.font90.blockquote.exercise[
```{r, echo = TRUE}
table4b %>% 
  pivot_longer(cols = c("1999", "2000"), names_to = "year", values_to = "population")
```
]

--

.font90[
See the help file for more information
]

???

notice that year is a character variable 
---
## Untidy Data

.blockquote.exercise[
```{r, echo = TRUE}
tidyr::table3 
```
]

???
How woud you describe the structure of the data?

--

- One column contains two variables. 

---
## Separating 

.font80[
The function `separate()` splits one column into multiple columns. 
As input it takes: 
- `col`, the column to be separated
- `into`, the names of the newly created columns
- `sep`, a regular expression which tells the function where to split (by default any non-alphanumeric character)
- `convert`, a logical value which controls whether or not to run `run type.convert()` on the new columns.
]

--

.code70[.blockquote.exercise[
```{r, echo = TRUE}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```
]]

???

why is sep not specified? 

---
## Untidy Data

.blockquote.exercise[
```{r, echo = TRUE}
tidyr::table5 
```
]

- One variable is spread over multiple columns.  

---
## Unite

The function `unite()` pastes multiple columns together. 
As input it takes: 
- `col`, the name of the new column
- `...`, the columns to be pasted together
- `sep`, the separator between values.

.blockquote.exercise[
```{r, echo = TRUE}
table5 %>% 
  unite(year, century, year, sep = "")
```
]

---
class: exercise_slide

## Exercises

<ol start="11"> 
<li>Compute the rate for <code>table2</code>, and <code>table4a</code> + <code>table4b</code>.  </li>

<li>Take the <code>tidyr::who</code> and <code>tidyr::who2</code> dataset.</li>
<ul>
<li>The who2 dataset is missing <code>iso2</code> and <code>iso3</code> codes. Create a lookup table using the original <code>tidyr::who</code> dataset that contains only the unique combinations of <code>country</code>, <code>iso2</code>, and <code>iso3</code>.</li>

<li>Use a join to add these codes to your who2 dataset.</li>

<li>Once joined, pivot the dataset so that all the tuberculosis measurement columns (from <code>sp_m_014</code> to <code>rel_f_65</code>) are moved into two columns: <code>category_code</code> and <code>count</code>.</li>

<li>Use <code>separate()</code> to split the <code>category_code</code> into three distinct variables: <code>method</code>, <code>sex</code>, and <code>age_group</code>.</li>

<li>The <code>age_group</code> codes are a bit messy (e.g., 014, 1524, 65). Use <code>mutate()</code> and <code>case_when()</code> to rename 014 to 00-14, 65 to 65+ and make all other groups use a hyphen (-) to seperate their limits.</li>

</ul>
<!--<li>  Recreate the plot given by the code below using <code>table2</code> instead of <code>table1</code>. </li>-->
</ol>

```{r, echo = F, eval = FALSE}
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

