---
title: "Exercise Solutions ggplot2"
author: 'Jens Klenke & Martin Schmelzer' 
output: 
  html_document:
   css: ["../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = F, echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 3, cache = T)
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css 
# https://github.com/gadenbuie/xaringanExtra
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"),
  rmarkdown::html_dependency_font_awesome() 
  )
```

```{r packages}
library(ggplot2)
library(readr)
library(here)
```


1. **Create a scatter plot of `mpg` vs. `wt` from the mtcars dataset. Map the `hp` variable to the color aesthetic.**

```{r}
# Mapping a continuous variable (hp) to color
ggplot(mtcars, aes(x = wt, y = mpg, color = hp)) +
  geom_point()
```


2. **Modify your previous plot. This time, map the `cyl` variable to the color aesthetic. Then, try mapping `cyl` to the shape aesthetic. If you encounter an error with shape, try wrapping the variable in `as.factor(cyl)`.**

```{r}
# Mapping cyl to color (treated as continuous by default)
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point()

# Mapping cyl to color and shape (treated as discrete/factor)
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl), shape = factor(cyl))) +
  geom_point()
```


3. **Create one more plot mapping a continuous variable to size and a categorical variable to color.**

```{r}
# Mixing continuous (disp for size) and categorical (am for color)
ggplot(mtcars, aes(x = wt, y = mpg, size = disp, color = factor(am))) +
  geom_point()
```

4. **What key observations do you note (in regards to how ggplot2 treats variables)?**

 - **Color Scales:** When a variable is numeric (like hp or cyl in its raw form), ggplot2 creates a continuous gradient (e.g., light blue to dark blue). When wrapped in factor(), it assigns distinct hues (discrete colors).

- **Shapes:** You cannot map a continuous variable to shape. ggplot2 will return an error because it is impossible to map a numerical spectrum to a finite set of symbols.

- **Legends:**
  - Continuous variables produce a color bar (a smooth transition).
  - Discrete variables produce a keyed list of individual points/colors.

Size: Mapping a variable to size works best for continuous data, as it creates a visual hierarchy of magnitude.



5. **Use the mtcars data set and plot mpg vs. hp. Add a smoothing line to the plot.**
    
```{r}
  ggplot(mtcars, aes(hp, mpg)) +
    geom_point() +
    geom_smooth()
```
    
6. **Add a smoothing function to the plot for each number of cylinders.**

```{r}
  ggplot(mtcars, aes(hp, mpg, color = factor(cyl))) +
    geom_point() +
    geom_smooth()
```
    
7. **Find out how to remove the confidence interval.**

```{r}
  ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
    geom_point() +
    geom_smooth(se = FALSE)
```
    
8. **Use a simple linear regression model and a quadratic regression model for smoothing.**

- Simple linear regression model

```{r}
    ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
      geom_point() +
      geom_smooth(method = lm)
```
  
- Quadratic regression model
  
```{r}
  ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
    geom_point() +
    geom_smooth(method = lm, formula = y ~ I(x + x^2))
```

9. **Can you visualize the "Women and Children First" rule? (Titanic Data on Moodle)**

```{r}
titanic <- readr::read_csv(here::here("SoSe_2026/ggplot2/titanic.csv"))
```

- Create a bar plot showing the total number of survivors vs. non-survivors.

```{r}
ggplot(titanic, aes(x = factor(survived))) + 
  geom_bar()
```

- Use the fill aesthetic to break down survival by Sex. Experiment with <code>position = "dodge"</code> vs <code>position = "stack"</code> — which makes the gender difference clearer?

```{r}
ggplot(titanic, aes(x = factor(survived), fill = sex)) + 
  geom_bar(position = "dodge")
```

- Use <code>facet_wrap(~ Pclass)</code> to see if your survival chances changed based on your ticket tier.

```{r}
ggplot(titanic, aes(x = factor(survived), fill = sex)) + 
  geom_bar(position = "dodge") +
  facet_wrap(~ pclass)
```
  
- Compare the age distribution between survivors and non-survivors.

```{r}
ggplot(titanic, aes(x = age, fill = factor(survived))) + 
  geom_density(alpha = 0.5)
```

10. **Try to answer the following questions about the `mpg` dataset (comes with ggplot2) using ggplot.**

- How are engine size and fuel economy related?

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ year)
# In general its a negative relationship:higher displacement (the bigger the engine) leads to less miles per gallon
```

- Do certain manufacturers care more about economy than others?

```{r}
ggplot(mpg, aes(x = hwy, fill = manufacturer)) +
  geom_density(alpha = 0.75)
# OR
ggplot(mpg, aes(x = reorder(manufacturer, hwy), y = hwy)) + 
  geom_boxplot() + 
  coord_flip()
```

- Has fuel economy improved in the last ten years?

```{r}
ggplot(mpg, aes(x = hwy, fill = factor(year))) +
  geom_density(alpha = 0.5)
```
    
11. **Take the two data sets economics and economics_long (both come with ggplot). Using each on of them: try plotting all the econonmic variables (unemploy, pop, psavert, etc.) over time on a single line plot with different colors.**
    
```{r}
  ggplot(economics, aes(x = date)) + 
    geom_line(aes(y = pop), col = "darkgrey") + 
    geom_line(aes(y = uempmed), col = "blue") + 
    geom_line(aes(y = pce), col = "orange") + 
    geom_line(aes(y = psavert), col = "red") + 
    geom_line(aes(y = unemploy), col = "green")
```

```{r}
  ggplot(economics_long, aes(x = date, y = value, color = variable)) + 
    geom_line()
```



`ggplot2` likes long data. If you want to plot multiple variables with color coding and legend, then you should transform your data set into a format similar to `economics_long`. See `tidyr` for switching between long and wide data formats.

12. **Reproduce the plot created by the following code with ggplot.**

```{r}
    plot(mtcars$mpg ~ mtcars$wt, xlab = "wt", ylab = "mpg", pch = 19, ylim = c(5, 35))
    mod <- lm(mpg ~ wt, data = mtcars)
    abline(mod, col = "red")
    wt_new <- seq(min(mtcars$wt), max(mtcars$wt), by = 0.05)
    conf_interval <- predict(mod, newdata = data.frame(wt = wt_new), 
                             interval = "confidence", level = 0.95)
    # setup vertrices of polygon (for shading the CI):
    p <- cbind(c(wt_new, rev(wt_new)), c(conf_interval[, 3], rev(conf_interval[, 2])))
    polygon(p, col = adjustcolor("steelblue", alpha.f = 0.5), )
    lines(wt_new, conf_interval[, 2], col = "steelblue", lty = 2)
    lines(wt_new, conf_interval[, 3], col = "steelblue", lty = 2)
```
    
    
```{r}
  ggplot(mtcars, aes(wt, mpg)) +
    geom_point(size = 2) +
    geom_smooth(method = lm, col = 'red', fill = 'steelblue') 
```
    
    



