---
title: "Advanced R for Econometricians"
subtitle: "`data.table`"
author: 
  - "Martin C. Arnold"
  - "Jens Klenke"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
class: title-slide title-datatable center middle
```{r moon_reader, include=FALSE, eval = FALSE}
# setup for moon reader
# copy it to console
options(servr.interval = 0.5)
xaringan::inf_mr()
```

```{r xaringanExtra_progress-bar, echo = FALSE}
xaringanExtra::use_progress_bar(color = "#004c93", location = "bottom")
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
options(htmltools.dir.version = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")


library(icons)
library(magrittr)

# pre-define icons
desktop <- icon_style(icons::fontawesome$solid$desktop, fill = "#004c93")
checker_green <- icons::icon_style(fontawesome$solid$check, fill = "green")
times_red <- icons::icon_style(fontawesome$solid$times, fill = "red")

```

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`

---
## What is a `data.table`?

The `data.table` package provides yet another alternative to data frames.
Similarly to `tibbles` the `data.table` class extends standard data frames. 

--

.blockquote.exercise[
```{r}
library(data.table)
DT <- data.table(a = rnorm(10), b = rnorm(10)) 
class(DT)
```
]

--

Compared to `tibbles` which provide only some convenience functionality over data frames,
`data.tables` are more like `tibbles` with integrated `dplyr`. 

--

### Some Interesting Links

- [Benchmarks](https://h2oai.github.io/db-benchmark/)
- [Comparison to `dplyr`](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/#chain-expressions)

---
## Creating a `data.table`

A data frame can be coerced to a `data.table` with `as.data.table()`. 

.blockquote.exercise[
```{r}
library(data.table)
df <- data.frame(a = rnorm(10), b = rnorm(10)) 
DT <- as.data.table(df)
class(DT)
```
]

--

However, the preferred way is to use `setDT()`.

.blockquote.exercise[
```{r}
setDT(df)
class(df)
```
]

--
<br>
What is the difference? 

---
## Modify by Reference

.blockquote.exercise[
```{r, results = 'hold'}
library(lobstr)
X <- data.frame(a = rnorm(10), b = rnorm(10)) 
obj_addr(X$a)
X <- as.data.table(X)
obj_addr(X$a)
```
]

--

.blockquote.exercise[
```{r, results = 'hold'}
X <- data.frame(a = rnorm(10), b = rnorm(10)) 
obj_addr(X$a)
setDT(X)
obj_addr(X$a)
```
]

---
## Why to use `data.table`

Compared to base `R` and the `tidyverse` the advantages of `data.table` are

- speed
- memory efficiency 
- compact syntax (which can be also a downside).

`data.table` can be used if the other options already exceeded their limits especially in terms of memory. 

--

The main functionality can be compared to `dplyr`

  - subsetting (`filter()` and `select()`)
  - updating (`mutate()`)
  - grouping and summarising (`group_by()` and `summarise()`)
  - merging  (`*_join()`)


---
## Syntax

With `dplyr` we use for each operation a single purpose function. 
With `data.table` (almost) all instructions are written in `[]`.

--


The general form of `data.table` syntax is:

.blockquote.exercise[
```{}
    DT[ i,  j,  by ] # + extra arguments
        |   |   |
        |   |    -------> grouped by what?
        |    -------> what to do?
         ---> on which rows?
```
]

--

Preparing a data.table:

.blockquote.exercise[
```{r}
library(ggplot2)
# setDT connot be used on data sets coming with packages 
DT <- as.data.table(diamonds)
```
]
---
## `i`: Filtering rows

- `i` is used for filtering rows. Note that no `$` is needed. 
--

.blockquote.exercise[
```{r}
DT[cut== "Fair" | cut == "Good"]
```
]

---
## i: keys

When setting the `key` attribute a `data.table` is sorted in memory using a [radix-sort](https://en.wikipedia.org/wiki/Radix_sort). This can be used for very fast lookup.

--

.blockquote.exercise[
```{r, eval = FALSE}
# one key column
setkey(DT,cut)
DT["Fair"]

# multiple key columns
setkey(DT,cut,color)
DT[list("Fair", "J")] 
```
]

<p style="margin: 30px;"></p>
--

`.(...)` is a convenient alias for `list(...)`. 

--

<p style="margin: 30px;"></p>

.blockquote.exercise[
```{r, eval = FALSE}
DT[.("Fair", "J")]
```
]

---
## Lookup

.font90[
A list or a `data.table` can be passed to `i` to run a lookup. (This is actually a join as we will see later.)

.blockquote.exercise[
```{r}
# Lookup table: 
LU <- data.table(color = c("E", "I") , clarity = c("SI2", "VS2")) 
```
]]

--

.font90[
<p style="margin: 20px;"></p>

For every row in LU, look up corresponding rows in DT,  using the variables specified in "on".

<p style="margin: 20px;"></p>

.blockquote.exercise[
```{r}
DT[LU, on = c("color", "clarity")]
```
]]
---
## j: Subsetting columns 

 `j` can be used for simple subsetting in the following ways:

- Returning a single vector.
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , price]
```
]

--

- Returning a `data.table` with only one column. 
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , .(price)]
```
]

--

- Returning `data.table` with multiple columns.
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , .(price, depth)]
DT[ , 1:2]
DT[ , c("price", "depth")]
```
]
---
## j: Subsetting columns 

- If the column names are stored in a variable use

--

.blockquote.exercise[
```{r, eval = FALSE}
cols <- c("price", "depth")
DT[ , ..cols]                # or
DT[ , cols, with = FALSE]
```
]
---
## j: Add and Remove Columns by Reference

- Add a variable. 
.blockquote.exercise[
```{r}
DT[ ,price_eur := price/0.91]
```
]

--

- Add multiple variables
.blockquote.exercise[
```{r}
# Here, we need to quote at the left hand side: 
DT[ ,c("x","y") := .(log(price), price - mean(price))]
```
]

--

- Remove one variable. 
.blockquote.exercise[
```{r}
DT[ , price_eur := NULL]
```
]

--

- Remove multiple variables. 
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , c("x", "price_eur") := NULL]
```
]

---
## Update Columns by Reference

- Simple update
.blockquote.exercise[
```{r}
DT[price > 2500, price := 2500]
```
]
--

<br>

- Updating multiple columns.
.blockquote.exercise[
```{r}
DT[price > 2500, c("price","depth") := .(log(price), sqrt(depth))]
```
]

---
## j: Evaluate Expressions

`j` can be used to evaluate expressions. Columns can be accessed as if they were in the global environment (compare to `dplyr::summarise()`).  

--

.blockquote.exercise[
```{r}
DT[ , mean(price) / sd(depth)]
```
]

The results are simply returned as a value which can be assigned to some variable. 

--

Several expressions can be used by putting them in a list. If a list is used, the result is returned as a `data.table`.

.blockquote.exercise[
```{r}
DT[ , .(mean_price = mean(price), sd_depth = sd(depth))]
```
]


---
## by: Grouping

The `by` argument allows operations by group (compare to `dplyr::group_by()`).

--

- Grouping by a single variable
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , .(mean_price = mean(price)), by = clarity]
```
]

--

- Grouping by multiple variables.
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , .(mean_price = mean(price)), by = .(clarity, color)]
```
]

--

- Grouping by expressions.
.blockquote.exercise[
```{r, eval = FALSE}
DT[ , mean(carat), by = list("Is price larger than 2300?" = price > 2300)]
```
]
---
## `keyby`

Additionally to grouping `keyby` runs `setkey()` on the by columns 	&rarr; results are sorted.

--

.blockquote.exercise[
```{r, echo = FALSE, eval = TRUE}
options(datatable.print.topn = 5, datatable.print.nrows = 5)
```

```{r, eval = TRUE}
# only the first and list five rows are printed
DT[ , mean(price), keyby = .(clarity, color)]
```
]


---
## .SD

Similar to `dplyr::across()` together with the predicate functions of the `dplyr` package we can aggregate multiple columns concisely using the special symbol `.SD`. 

If nothing else is specified, `.SD` corresponds to all columns in the `data.table`.
We can use this together with `lapply` to aggregate all columns using the same function. 

--

.blockquote.exercise[
```{r}
numeric_DT <- data.table(a = rnorm(10), b = rnorm(10), c = rnorm(10))
numeric_DT[ ,lapply(.SD, mean)]
```
]

--

We can however change `.SD` to only contain a subset of all columns using `.SDcols`.

.blockquote.exercise[
```{r}
num_cols <- names(DT)[sapply(DT, is.numeric)] 
DT[ , lapply(.SD, mean), .SDcols = num_cols ]
```
]


---
## Joins

.blockquote.exercise[
```{r}
library(dplyr)
members <- as.data.table(band_members)
instruments <- as.data.table(band_instruments)
```
]

--

- Left join (`members` is the left table )
.blockquote.exercise[
```{r}
instruments[members, on = "name"]
```
]

--

- Left join by reference
.blockquote.exercise[
```{r}
members[instruments, band := band, on = "name"]
```
]
---
## Joins
- Inner join
.blockquote.exercise[
```{r}
instruments[members, on = "name", nomatch = 0]
```
]

--

- A full join cannot be achieved with standard `data.table` syntax. Use `data.table::merge()`.
.blockquote.exercise[
```{r}
merge(members, instruments, all = TRUE)
```
]
---
## Joins
- Anti-Join
.blockquote.exercise[
```{r}
merge(members, instruments, all = TRUE)[is.na(plays)]
```
]
---
## Joins: Overview
<br>

.font90[
```{r, echo = FALSE, results='asis'}
data.frame("Join type" = c("INNER", "LEFT", "RIGHT", "FULL", "ANTI"),
           "DT" = c("X[Y, nomatch=0]", "X[Y]", "Y[X]", "-", "-"),
           "data.table::merge()" = c("merge(X, Y, all=FALSE)", "merge(X, Y, all.x=TRUE)", "merge(X, Y, all.y=TRUE)", "merge(X, Y, all=TRUE)", "merge(X, Y, all=TRUE)[is.na(...)]"),
           "dplyr" = c("inner_join(X,Y)", "left_join(X,Y)", "right_join(X,Y)", "full_join(X,Y)", "anti_join(X,Y)"),
           check.names = FALSE) %>% knitr::kable(., format = "html")
```
]

---
## Rolling Join 

.blockquote.exercise[
```{r}
DT <- data.table(t = c(1.2, 2.8, 4.5, 10), value = c(3, 8, 4, 2), id_DT = 1:4) 
LU <- data.table(id_LU = 1:3, t = c(3.3, 1.0, 9.0)) 
DT[LU, on = "t", roll = TRUE]
```
]

--

- Looks for closest value of `t` in `DT` which is smaller than `t` in `LU`.

---
## Other Useful Functions from the `data.table` package

-  `fread()`, `fwrite()`: very fast data reading and writing
-  `rbindlist()`: Creates a `data.table` from a list of `data.tables`
-  `rowid():` Creates groupwise row ids.
-  `fsetdiff(DT1,DT2)`, `funion(DT1,DT2)`, ...: Fast set operations for `data.tables`
-  `uniqueN(cut)`: Number of unique elements
-  `shift()`: For lead and lag

---
class: exercise_slide
## Exercises 

1. Repeat the exercises from the chapter on `dplyr` using `data.table`. 
2. Compare the speed of reading in the Boston crime dataset with `read.csv()` and `fread()`. Use the 
   function `system.time()` to find out how long it takes. 













