---
title: "Advanced `R` for Econometricians (Summer 2025)"
subtitle: "Object Oriented `R` Programming --- Solutions to Exercises"
author: 'Martin Arnold, Martin Schmelzer' 
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(sloop)
```

# Slide 18

### 1. 

Because of S3’s `generic.class` naming scheme, both functions may initially look similar, while they are in fact unrelated.

`t.test()` is a generic function that performs a t-test. `t.data.frame()` is a method that gets called by the generic `t()` to transpose dataframe input.
Due to `R`’s S3 dispatch rules, `t.test()` would also get called when `t()` is applied to an object of class test

### 2.

We obtain an object of class `ecdf` (short for empirical cumulative distribution function) with the superclasses `stepfun` and `function`. The `ecdf` object is built on the base type `closure` (a function). The expression which was used to create it (`rpois(100, 10)`), is stored in the `call` attribute.

### 3.

The code returns a `table` object, which is built upon the integer type. The attribute `dimnames` is used to name the elements of the integer vector.


# Slide 36 

### 1.

We see that `t.test()` is a generic because it calls `UseMethod()`:

```{r}
t.test

# or simply call
sloop::ftype(t.test)
```


Interestingly, `R` also provides helpers, which list functions that look like methods, but in fact are not:

```{r}
tools::nonS3methods("stats")
```
When we create an object with class test, `t()` dispatches to the `t.default()` method. This is because `UseMethod()` simply searches for functions named `paste0("generic", ".", c(class(x), "default"))`.

```{r}
x <- structure(1:10, class = "test")

t(x)
```

However, in older versions of `R` (pre 4.0.0) this behaviour was slightly different. Instead of dispatching to the `t.default()` method, the `t.test()` generic was erroneously treated as a method of `t()` which then dispatched to `t.test.default()` or (if defined) to `t.test.test()`.

### 2.

This is a simple application of `sloop::s3_methods_class()`:

```{r}
s3_methods_class("table")
```

Interestingly, the `table` class has a number of methods designed to help plotting with base graphics.


### 3.

```{r}
prime <- function(x = integer()) {
  validate_prime(new_prime(x))
}

validate_prime <- function(x = list()) {
  sapply(
    unique(unlist(x)), function(z) {
      if(!all(z==2 || z %% 2:(z-1) > 0)) {
        stop('Input contains non-prime number(s)!', call. = F)
      }
    } 
  )
  x
}

new_prime <- function(x = integer()) {
  stopifnot(is.integer(x))
  x <- list(x)
  class(x) <- 'prime'
  x
}
```

### 3.1

```{r}
summary.prime <- function(x) {
  if(class(x)  == "prime") {
    x <- unlist(x)
    out <- list(
        name = quote(x),
        n = length(x),
        min = min(x),
        max = max(x)
        )
    class(out) <- "summary.prime"
    return(out)
  } else {
    message("Object not of class prime!")
  }
}

my_primes <- new_prime(c(3L, 5L, 7L))

summary(my_primes)
```

### 3.2

```{r}
print.summary.prime <- function(x) {
  if(class(x)  == "summary.prime") {
    cat(
      "summary for", x$name, ":\n\n",
      x$n, "prime numbers between", x$min, "to", x$max
    )
  } else {
    message("Object not of class summary.prime!")
  }
}

# now with print method
summary(my_primes)
```




