---
title: "Advanced R for Econometricians"
subtitle: "Databases using R"
author: 
  - "Martin Arnold"
  - "Martin Schmelzer"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/navigate_home.js", "../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
class: title-slide title-sqlite center middle
```{r icons, include=FALSE}
library(icons)

# pre-define icons
desktop <- icons::icon_style(icons::fontawesome$solid$desktop, fill = "#004c93")
checker_green <- icons::icon_style(icons::fontawesome$solid$check, fill = "green")
times_red <- icons::icon_style(icons::fontawesome$solid$times, fill = "red")

library(keyring)
```

```{r xaringanExtra_progress-bar, echo = FALSE}
xaringanExtra::use_progress_bar(color = "#004c93", location = "bottom")
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r, include = FALSE, cache = FALSE}
library(knitr)
knitr::opts_chunk$set(eval = FALSE)

default_source_hook <- knit_hooks$get("source")
knit_hooks$set(
  source = function(x, options) {
    if(is.null(options$table))
      default_source_hook(x, options)
    else {
      eval(parse(text = x)) %>%
        kable("html") %>%
        kable_styling("hover", full_width = F)
    }
  }
)
```

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`

---
## Memory Limitations of R

- The amount of data you can work with in R is mainly constraint by your computers memory. 
- In times where data sets with several 100 GB or even TB become more and more common,
  loading all data into memory is usually not feasable. 
- Often only a subset or an aggregation of the data is required for the analysis which requires much less memory space.
- How to subset or aggregate data? &rarr; Use a database!

---
## (Relational) Databases

.font90[
- A database is a place where data is stored and can be manipulated.
- A relational database aims to reduce redundancies by spreading the data over multiple tables and
  relate them to each other. 
]

--

.font90[.blockquote.exercise[

#### `r desktop` Example: Redundent Data

```{r, echo = FALSE, eval = TRUE, results = 'asis', message = FALSE}
library(dplyr)
set.seed(123)
cust_names<- c("17" = "Peter", "34" = "Kelly", "108" = "Aleta")
cust_address<- c("17" = "Carnaby  St.", "34" = "Downing Rd.", "108" = "Berwick St.")
sells <- tibble(store_id = rep(1:2, each = 4), 
                    store_location   = rep(c("Baker St.", "Abbey Road"), each = 4),
                    store_mngr   = rep(c("John", "George"), each = 4),
                    cust_id      = sample(c(34, 17, 108, 17, 108, 17, 34, 17), 8, replace = TRUE),
                    cust_address = cust_address[as.character(cust_id)],
                    cust_name    = cust_names[as.character(cust_id)],
                    date = sample(seq(as.Date("2019/1/1"), as.Date("2019/3/1"), 
                               by = "day"), 8),
                    total = sample(10:20, 8, replace = TRUE)
                )

knitr::kable(sells, format = "html")
```

<br>
]]
---
#### Example of Non-Redundent Data 

- The same information would be stored in a relational database like this: 

--


```{r, eval = TRUE, echo = FALSE}
store <- tibble(store_id =  1:2, 
                store_location = c("Baker St.", "Abbey Road"),
                store_mngr   = c("John", "George")
                )

customer <- tibble(cust_id = names(cust_names),
                   cust_names = cust_names,
                   cust_address = cust_address)

sells_rel <- sells %>% select(-store_location, -store_mngr, -cust_address, -cust_name) 
```

.pull-left[
```{r, echo = FALSE, eval = TRUE}
knitr::kable(sells_rel, format = "html", caption = "Sell Records")
```
]
.pull-right[
```{r, echo = FALSE, eval = TRUE}
knitr::kable(store, format = "html", caption = "Store")
```
<br>
```{r, echo = FALSE, eval = TRUE}
knitr::kable(customer, format = "html", caption = "Customer")
```
]

<br>

--
- The relationship between the tables is defined by the `id` columns (called keys).



---
## R and Databases

- `dplyr` also works with remote on-disk data stored in relational databases.
- For this to work you need 

--

.blockquote.exercise[
```{r, eval = FALSE}
install.packages("dbplyr")
```
]

--

- You’ll also need to install a `DBI` backend package. The `DBI` package provides a common interface that allows `dplyr` to work with many different databases using the same code. 

--

- Some commonly used backends are:
    - `RMariaDB` connects to MySQL and MariaDB.
    - `RPostgreSQL` connects to Postgres and Redshift.
    - `RSQLite` embeds a SQLite database.
    - `duckdb` embeds a fast, in-process analytical db optimized for large data and complex queries.
    - `bigrquery` connects to Google’s BigQuery.

---
## Connect to a Database

.font80[
- To connect to a database authentication information such as 
  username and password are required (the database admin should provide those). 
- You also need to specify a driver that makes communication with the database possible.
- ~~There is a remote MySQL database hosted on the web which we have access to:~~
]

--

.font80[.blockquote.exercise[

#### `r desktop` Setup the connection 

```{r, eval = FALSE}
library(dbplyr)
library(DBI)
con <- DBI::dbConnect(
          drv = RMariaDB::MariaDB(), # MariaDB driver works for MySQL as well
          username = keyring::key_list('sql_try')[1,2],
          dbname   = keyring::key_list('sql_try')[1,2], 
          host     = "sql11.freesqldatabase.com",
          password = keyring::key_get('sql_try',
                                      keyring::key_list('sql_try')[1,2]),
          port     = "3306"
          )
```
]]

```{r, eval = FALSE, echo = FALSE}
library(dplyr)
con <- DBI::dbConnect(
               drv = RMariaDB::MariaDB(), # MariaDB driver works for MySQL as well
               username = 'sql11460561',
               dbname   = keyring::key_list('sql_try')[1,2], 
               host     = 'sql11.freesqldatabase.com',
               password = keyring::key_get('sql_try',
                                           keyring::key_list('sql_try')[1,2]),
               port     = '3306'
               )
```









---

.font90[
## Set up an SQLite Database

- SQLite is a very light weighted Database Management System 
- An SQLite database is contained in a single file. There is no need for a server. 
- Everything you need is to install the `RSQLite` packages. 
]

--

.font90[
#### Create a SQLite Database
- A new SQLite database is automatically created if you call `DBI::dbConnect()`
with argument `dbname = "path_to_database"` where `"path_to_database"` is a path to 
a non-existing file. 
- In case `"path_to_database"` is an existing SQLite database it 
connects to that database. 
]

--

.font90.blockquote.exercise[
#### `r desktop` Example 

```{r, echo = TRUE, eval = FALSE}
library(dplyr)
con <- DBI::dbConnect(
  drv = RSQLite::SQLite(),
  dbname = "local_db.sqlite"
)
```
]

???

SQLite -> SQLite is a C-language library that implements a small, fast, self-contained, high-reliability, full-featured, SQL database engine.



---
## Write Data to the Database
.font80[
- When the connection is established we can use the object `con` to communicate with the database.
- We add the data frame `mtcars` as a new table to the database.
]
--

.font90.blockquote.exercise[
#### `r desktop` Example 
```{r, eval = FALSE}
DBI::dbWriteTable(con,
                  name = "tbl_mtcars", 
                  value = mtcars)
```
]

--
.font80[
- To add data to an existing table we use the additional argument `append = TRUE`.
]

--

.font90.blockquote.exercise[
#### `r desktop` Example 

```{r, eval = FALSE}
new_car <- data.frame(mpg = 1, cyl = 16, disp = 800, hp = 210, 
                      drat = 5, wt = 6, qsec = 2, vs = 1, am = 0)
rownames(new_car) <- "Super Car 500"

DBI::dbWriteTable(con, 
                  name = "tbl_mtcars", 
                  value = new_car, 
                  append = TRUE)

DBI::dbDisconnect(con)
```
]

???
with name `mtcars_db` 



---
## Write Data to the Database


.font90.blockquote.exercise[
#### `r desktop` Example 

```{r, echo = TRUE, eval = FALSE}
dbWriteTable(con, "tbl_gapminder", gapminder::gapminder)
dbWriteTable(con, "tbl_country_codes", gapminder::country_codes)
```
]


---
## Access Data

- The function `DBI::dbListTables()` tells us which tables are in the database.

.blockquote.exercise[

```{r, eval = FALSE}
DBI::dbListTables(con)
```
]

--

- With `dplyr::tbl()` we can create an object which is a reference to a table in the database. Most `dplyr` functions work on this object as it was as a data frame (or `tibble`).

--

.blockquote.exercise[

#### `r desktop` Example 
```{r, eval = FALSE}
gap_tbl <- tbl(con, "tbl_gapminder")
class(gap_tbl)
```
]
---
## Use Standard `dplyr`  Functions on the Database

.font90.blockquote.exercise[

#### `r desktop` Example 
```{r, message=FALSE, warning=FALSE, eval = FALSE}
gap_avg_tbl <- gap_tbl %>%
  group_by(continent) %>% 
  select_if(is.numeric) %>% 
  select(-year) %>%
  summarise_all(mean)
```

]

--

.font90[
- Even if it seems as if `gap_avg_tbl` is a data frame it still is only a set of instructions to be performed by the database. 
- `show_query()` tells us which instructions are sent to the database. 
]

--

.font90.blockquote.exercise[

#### `r desktop` Example 
```{r, warning=FALSE, eval = FALSE}
show_query(gap_avg_tbl)
```
]

???


- select_if superseded
- but dbplyr **do not** support predicate functions, i.e. you can not use where(is.numeric) to select all numeric variables.



---
## Use Standard `dplyr` Functions on the Database

.font90[
When working with databases, `dplyr` tries to be as lazy as possible:

- It never pulls data into R unless you explicitly ask for it.
- It delays doing any work until the last possible moment: it collects everything you want to do and then sends it to the database in one step.
- `collect()` actually pulls the data into R.  
]

--


.font90.blockquote.exercise[
#### `r desktop` Example 

```{r, eval = FALSE}
gap_avg <- gap_avg_tbl %>% collect(n = Inf)
print(gap_avg)
```
]
---
## Disconnect

- If the connection to the database is no longer needed you should disconnect to free system resources. 

--

.blockquote.exercise[
```{r, eval = FALSE}
DBI::dbDisconnect(con)
rm(con)
```
]




---
class: exercise_slide
## Exercises

1. Set up a new SQLite database containing the Boston Crime data. 
2. Without retrieving the data to R let the database compute the percentage of incidents involving a shooting. 












