---
title: "Rcpp --- Solutions to Exercises"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## `Rcpp` &mdash; Exercises Part 1

- Find out why the following code gives a compile error:
    ```{Rcpp, eval=F}
    #include<Rcpp.h>
    // [[Rcpp::plugins("cpp11")]]
    
    NumericVector x{1, 2, 3, 4, 5};
    IntegerVector id{1};
    
    double y = x[id]; // produces compile error
    ```

    **Solution:**
    
    `x[id]` returns a subview class of `Rcpp::Vector` which is *not* a double. It is important to note that coercion to another type is not as easily done as in `R`!

- Benchmark the functions below against each other for `x<-rnorm(1e2)`, `x<-rnorm(1e4)` and `x<-rnorm(1e6)`. Comment on the results.
    ```{r, echo = F}
    cppFunction('
        NumericVector test_clone_return(NumericVector A) {
      NumericVector B = clone(A);
      B[1] = 0.5;
      return B;
    }
    ')
    cppFunction('
    NumericVector test_reference_return(NumericVector A) {
      A[1] = 0.5;
      return A;
    }
    ')
    ```
    
    ```{Rcpp, eval=F}
    NumericVector test_clone_return(NumericVector A) {
      NumericVector B = clone(A);
      B[1] = 0.5;
      return B;
    }
    ```
    ```{Rcpp, eval=F}
    NumericVector test_reference_return(NumericVector A) {
      A[1] = 0.5;
      return A;
    }
    ```

  **Solution:**

    ```{r}
    bench::press(
      A = c(1e2, 1e4, 1e6),
      {
        A <- rnorm(A)  
        bench::mark(
          test_clone_return(A),
          test_reference_return(A), 
          check = F,
          relative = T
        )
      }
    )
    ```
    
Obviously, cloning (that is, copying on function call) is a bad idea!

## Slide 40: `Rcpp` -- STL Algorithms

The second appearance of `x.begin()` refers to the beginning of the output range, i.e. the first element of `x`.

 
## Slide 44 `Rcpp` &mdash; Exercises Part 2 

1. Note that `f(n)` returns the $n^th$ Fibonacci number. Here are two `Rcpp` approaches. `f_rec_cpp()` is a one-to-one implementation of the recursive `R` function `f()`. `f_cpp()` uses a loop for the computation.

    ```{Rcpp, eval = F}
    // [[Rcpp::export]]
    int f_rec_cpp(const int& n) {
      if (n < 3) return(n); 
      return (f_rec_cpp(n - 1)) + f_rec_cpp(n - 2);
    }    
        
    // [[Rcpp::export]]
    IntegerVector f_cpp(const int& n) {
      IntegerVector x(n);
      x[0] = 1;
      x[1] = 2;
      for (int i = 2; i < n; i++){
        x[i] = x[i-2] + x[i-1];
      }
      return(tail(x, 1));
    }
    ```

    ```{R, echo = F, eval = T}
    f <- function(n) {
      if(n < 3) return(n)
      return(f(n-1) + f(n-2))
    }
    
    Rcpp::cppFunction("
    int f_rec_cpp(const int& n) {
      if (n < 3) return(n); 
      return (f_rec_cpp(n - 1)) + f_rec_cpp(n - 2);
    }")
    
    Rcpp::cppFunction("
    IntegerVector f_cpp(const int& n) {
      IntegerVector x(n);
      x[0] = 1;
      x[1] = 2;
      for (int i = 2; i < n; i++){
        x[i] = x[i-2] + x[i-1];
      }
      return(tail(x, 1));
    }")
    ```

    Let's microbenchmark these functions:
  
    ```{r, cache = T, eval = T}
    bench::mark(
      f = f(20),
      f_rec_cpp(20),
      f_cpp = f_cpp(20),
      relative = T
    )
    ```

    The overhead of recursive function calling in `R` is much bigger than in `C++`. The `Rcpp` implementation `f_rec_cpp()` already gives an impressive speedup. However, using a `for` loop in `f_cpp` is even faster!
    
2. 

    ```{R, echo = F, eval = T}
    Rcpp::cppFunction("
      NumericVector nth_partial_sort(const NumericVector& x, const int& nth) {
        NumericVector y = clone(x);
        std::nth_element(y.begin(), y.begin()+nth, y.end());
        std::sort(y.begin(), y.begin()+nth);
        return y;
      }")
    ```
    
    ```{R, eval = T}
    set.seed(1234)
    
    # let's generate a large vector to be (partially) sorted
    v <- rnorm(1e5)
    n <- 1e4
    
    # benchmark n^th partial sort
    bench::mark(
      nth_partial_sort(v, n),
      sort(v, partial = 1:n), 
      check = F,
      relative = T
    )
    ```

3. Here are very basic `Rcpp` versions for vector input:

    ```{R}
    Rcpp::cppFunction('
    bool all_cpp(const LogicalVector& x) {
      int n = x.size();
      
      for(size_t i = 0; i < n; i++) {
        if(x[i] != true) return false;
      }
    
      return true;
    }
    ')
    
    Rcpp::cppFunction('
    NumericVector range_cpp(const NumericVector& x) {
      NumericVector out = {min(x), max(x)};
      return out;
    }
    ')
    
    Rcpp::cppFunction('
    double var_cpp(const NumericVector& x) {
       int n = x.size() - 1;
       return 1./n * sum(pow(x - mean(x), 2));
    }
    ')
    
    ```









    