---
title: "More Topics"
author: "Martin Schmelzer"
date: "2025-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Topic 9: Birth Country Encryption

### Key Concepts

tidyverse, webscraping, functional programming, shiny

### Short Description 

Inferring the country of birth from place of birth and auxiliary features.

### Tasks:

- Obtain reference data on locations around the world (possibly also historical).

- Try to fill in the missing values for the "birth_country" variable using the reference data and auxiliary features (1st citizenship, 2nd citizenship). You are free to choose the method.
- Build a shiny app with the following elements/features:
  - An upload button to load a file containing unencrypted birth countries
  - The process can be started through a button.
  - Display of the data/results in an interactive table
  - Output the success rate for 100% hits (e.g. unique city name + citizenship fits the bill -> country x)
  - Visualization of the frequency of birth countries
  
  
```{r, message=F, warning=F,  echo=F}
library(tidyverse)
df <-readr::read_csv2(here::here("SoSe_2025/Topics/birthdata.csv"))
head(df)
```

