---
title: "Solution"
author: "Alexander Gerber & Martin Arnold"
date: "11/11/2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Experiment with geom_point() using the mtcars data set. Try out different aesthetics with different variables. What do you note?

    Plot weight against horsepower:

    ```{r}
    library(ggplot2)
    ggplot(mtcars, aes(x = hp, y = wt)) +
      geom_point()
    ```

    Plot weight against color the points according to gear:

    ```{r}
    ggplot(mtcars, aes(x = hp, y = wt, color = gear)) +
      geom_point()
    ```

    Plot weight against color the points according to gear as a categorical variable:

    ```{r}
    ggplot(mtcars, aes(x = hp, y = wt, color = forcats::as_factor(gear))) +
      geom_point()
    ```

    Use gear on the y axis.

    ```{r}
    ggplot(mtcars, aes(x = hp, y = gear)) +
      geom_point()
    ```
    
2. Use the mtcars data set and plot mpg vs. hp. Add a smoothing line to the plot.
    
    ```{r}
    ggplot(mtcars, aes(hp, mpg)) +
      geom_point() +
      geom_smooth()
    ```
    
3. Add for each number of cylinders a linear smoothing function to the plot.

    ```{r}
    ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
      geom_point() +
      geom_smooth()
    ```
    
4. Find out how to remove the confidence interval.

    ```{r}
    ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
      geom_point() +
      geom_smooth(se = FALSE)
    ```
    
5. Use a simple linear regression model and a quadratic regression model for smoothing.

    Simple linear regression model

    ```{r}
    ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
      geom_point() +
      geom_smooth(method = lm)
    ```
    
    Quadratic regression model
    
    ```{r}
    ggplot(mtcars, aes(hp, mpg, color = as.factor(cyl))) +
      geom_point() +
      geom_smooth(method = lm, formula = y ~ I(x^2))
    ```
    
7. Download the Titanic data set from Moodle and

    Use a bar plot to show how many people survived the Titanic compared to those who didn't.

    ```{r}
    titanic <- readr::read_csv("titanic.csv")
    ggplot(titanic, aes(x = survived)) +
      geom_bar()
    ```

    Add a color coding to the previous plot to visualize the differences between the passengers gender.

    ```{r}
    ggplot(titanic, aes(x = survived, fill = sex)) +
      geom_bar()
    ```

    Split the previous plot into three plots based on the passengers class.

    ```{r}
    ggplot(titanic, aes(x = survived, fill = sex)) +
      geom_bar() +
      facet_grid(.~pclass)
    ```

    Compare the age distribution between the survivors and non-survivors.

    ```{r}
    ggplot(titanic, aes(x = age, colour = survived, fill = survived)) +
      geom_density(alpha = 0.5)
    ```

8. Try to answer the following questions about the mpg dataset (comes with ggplot2) using ggplot.

    How are engine size and fuel economy related?

    ```{r}
    ?mpg
    ggplot(mpg, aes(x = displ, y = hwy)) +
      geom_point()
    ```

    Do certain manufacturers care more about economy than others?

    ```{r}
    ?mpg
    ggplot(mpg, aes(x = displ, y = hwy, color = manufacturer)) +
      geom_point()
    ```

    Has fuel economy improved in the last ten years?

    ```{r}
    ggplot(mpg, aes(x = year, y = hwy, color = manufacturer)) +
      geom_point()
    ```
    
9. Compare the two data sets economics and economics_long (both come with ggplot) with respect to the ease of use when working with ggplot.

    ```{r}
    head(economics, 20)
    ```
    
    ```{r}
    head(economics_long, 20)
    ```
    
    The economics dataset should be preferred over economics_long because of its tidyness. Some plots in ggplot2 require untidy data but most of the time a call of the gather() function will be sufficient to solve the problems.
    

10. Reproduce the plot created by the following code with ggplot.

    ```{r}
    plot(mtcars$mpg ~ mtcars$wt, xlab = "wt", ylab = "mpg", pch = 19, ylim = c(5, 35))
    mod <- lm(mpg ~ wt, data = mtcars)
    abline(mod, col = "red")
    wt_new <- seq(min(mtcars$wt), max(mtcars$wt), by = 0.05)
    conf_interval <- predict(mod, newdata = data.frame(wt = wt_new), 
                             interval = "confidence", level = 0.95)
    # setup vertrices of polygon (for shading the CI):
    p <- cbind(c(wt_new, rev(wt_new)), c(conf_interval[, 3], rev(conf_interval[, 2])))
    polygon(p, col = adjustcolor("steelblue", alpha.f = 0.5), )
    lines(wt_new, conf_interval[, 2], col = "steelblue", lty = 2)
    lines(wt_new, conf_interval[, 3], col = "steelblue", lty = 2)
    ```
    
    
    ```{r}
    ggplot(mtcars, aes(wt, mpg)) +
      geom_point(size = 2) +
      geom_smooth(method = lm, col = 'red', fill = 'steelblue') +
      theme_linedraw()
    ```
    
    