```{r}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, eval = FALSE)


```


1. Load the nycflights13::flights data set. Does it come formatted as a tibble?

```{r}
library(tidyverse)
library(nycflights13)
data("flights")
flights
```

2. Find all flights that
    - had an arrival delay of two or more hours.
    ```{r}
    flights %>% filter(arr_delay >= 120)
    ```
        
   - had an arrival delay of two or more hours.
    ```{r}
    flights %>% filter(dest %in% c("IAH", "HOU"))
    ```
   - were delayed by at least an hour, but made up over 30 minutes in flight.
    ```{r}
    flights %>% filter(dep_delay > 60 & arr_delay <= dep_delay - 30)
    ```
   - departed between midnight and 6 am.
    ```{r}
    flights %>% filter(between(sched_dep_time, 0, 600))
    ```
   
3. Which flights took the longest? Which traveled the shortest? 
    ```{r}
    flights %>% slice(which.max(air_time))
    flights %>% slice(which.min(air_time))
    ```

4. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay`.

```{r}
selection <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
flights[ ,selection]

flights %>% select(selection)
```

5. Create a new variable `first_flight` which takes the value `TRUE` if it is the first flight of the day.
```{r}
# First way
flights <- flights %>% 
  mutate(first_flight = if_else(day - lag(day) != 0 | is.na(day - lag(day)), TRUE, FALSE )) 

# flights <- flights %>% 
#   group_by(year, month, day) %>% 
#   mutate(first_flight = if_else(row_number() == 1, TRUE, FALSE )) %>%
#   select(first_flight)
```

6.Find out in which district the most incident where reported. 
```{r}
crime <- as_tibble(read.csv(here::here("datasets", "BostonCrime.csv")))
crime %>% 
  group_by(DISTRICT) %>%
  summarise(N = n()) %>% 
  arrange(desc(N))
```

7.Find out in which district the most incident where reported. 

```{r}
crime %>% 
  filter(OFFENSE_CODE_GROUP == "Drug Violation") %>% 
  group_by(DISTRICT) %>% 
  summarise(N = n()) %>% 
  arrange(desc(N)) 
```

8. What is the most dangerous hour of the week?

```{r}
crime %>% 
  filter(OFFENSE_CODE_GROUP == "Homicide") %>% 
  group_by(HOUR, DAY_OF_WEEK) %>%
  summarise(N = n()) %>% 
  arrange(desc(N))
```

10. Calculate for each district the proportion of motor vehicle accidents to all reported incidents.

```{r}
crime %>% 
  group_by(DISTRICT) %>%  
  transmute(prop_accidents = sum(OFFENSE_CODE_GROUP == "Motor Vehicle Accident Response") / n()  )
  
```

11. Sort the data set chronologically by year and month. Then take the first 10.000 observations and compute the proportion of incidents where a shooting was involved for each district. Now take a random sample of size 10.000. Is there a noteworthy difference?

```{r}
set.seed(1245)

crime %>% 
  arrange(desc(YEAR), desc(MONTH)) %>%
  slice(1:10000) %>% 
  summarise(shooting = mean(SHOOTING == "Y"))

crime %>% 
  sample_n(10000) %>% 
  summarise(shooting = mean(SHOOTING == "Y"))
```

12. Compute the rate for `table2`, and `table4a` + `table4b`.

```{r}
# Using table 2. Adding the computed data to the table would be inconvenient.
table2[table2$type == "cases", "count"] / table2[table2$type == "population", "count"]
# Using table4a + `table4b
table4a[ ,-1] / table4b[ ,-1]
```


13. Recreate the plot given by the code below using `table2` instead of `table1`.

```{r}
## Using Table 1
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

## Using Table 2
ggplot(table2 %>% filter(type == "cases"), aes(x = year, y = count)) +
    geom_line(aes( group = country)) +
    geom_point(aes(color = country))

```










   
   
        