---
title: "Exercise Solutions Introduction"
author: 'Jens Klenke' 
output: 
  html_document:
   css: ["../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css 
# https://github.com/gadenbuie/xaringanExtra
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"),
  rmarkdown::html_dependency_font_awesome() 
  )
```

## Exercise 

1. Combine all elements of `df_list` to one data frame. The result should be only a single line of code.
  - `do.call(rbind.data.frame, df_list)`
    - `do.call` pass elements of `df_list` as arguments to `rbind`. It's equivalent of `rbind(df_list[[1]], df_list[[2]], df_list[[3]], ....., df_list[[length of df_list]])`. 
2. Fix each of the following common data frame subsetting errors:
  - `mtcars[mtcars$cyl == 4, ]`
  - `mtcars[-c(1:4), ]`
  - `mtcars[mtcars$cyl <= 5, ]`
  - `mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]`
3. What does `df[is.na(df)] <- 0` do
  - replaces all `NA`s with `0` but keeps the dimensions. 
4. How would you randomly permute the columns of a data frame?
  - `df[sample(1:nrow(df)), ]`
5. Write a `while` and a `repeat` loop doing the same as
```{r}
for(i in LETTERS[1:10]){
  print(i)
}

# while 
i <- 0
while (i < 10) {
  i = i + 1
  print(LETTERS[i])
  
}
# alternative 
i <- 1
while (i < 11) {
  print(LETTERS[i])
  i = i + 1 
}

# repeat
i <- 0
repeat{
  i = i + 1
  print(LETTERS[i])
  if (i > 9) {
    break
  }
  
}

# alternative
i <- 1
repeat{
  print(LETTERS[i])
  i = i + 1 
  if (i > 10) {
    break
  }
}
```
6. Write a `for` loop doing the same as

```{r}
count <- 0
repeat{
  x     <- sample(1:6, 1)
  count <- count + 1 
  if(x == 6) break
}
print(count)

for (i in 1:100) {
  x     <- sample(1:6, 1)
  count <- i
  if(x == 6) break
}
print(count)
```
  - note that $n$ has to be high enough
7. What is the problem here?
  - Curly braces are open and closed before the statement is passed to the function
8. What is the problem here? Can you debug it?
  - `df` will not be passed to the final call and `fast_lm` is the wrong call
  ```{r}
  
df <- mtcars

lin_mod <- function(df){
  lm(y ~ . ,data = df)
}

simple_lin_mod <- function(data, y, x){
  df <- data[ ,c(y, x)]
  names(df) <- c("y", "x")
  lin_mod(df)
}

simple_lin_mod(df, "mpg", "cyl")
```

