---
title: "Exercise Solutions Introduction"
author: 'Jens Klenke' 
output: 
  html_document:
   css: ["../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r xaringanExtra-clipboard_2, echo=FALSE}
# copy button styles mainly in ude.css 
# https://github.com/gadenbuie/xaringanExtra
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #00ff00\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"),
  rmarkdown::html_dependency_font_awesome() 
  )
```

#### Slide 13

1. list can store different data types
2. data frames can store different values with different types
3. class()
4. is.character()

#### Slide 14

1. 
  - `[]` = always returns object of same class 
  - `[[]]` = can extract one element from list or data frame
  - `list_a <- list(a = 2,   b = "hallo")`
    - `list_a["a"]`
    - `list_a[["a"]]`
    - `list_a$a`

2. 
  - `[` selects sub-lists. It always returns a list; if you use it with a single positive integer, it returns a list of length one. 
  - `[[` selects an element within a list. 
  - `$` is a convenient shorthand: `x$y` is equivalent to `x[["y"]]`.

3. 
  - Use `drop = FALSE` if you are subsetting a matrix, array, or data frame and you want to preserve the original dimensions. You should almost always use it when subsetting inside a function.


#### Slide 15

1.
  - `if()` works with scalars; `ifelse()` works with vectors.
2.
  - when `x` is `TRUE`, `y` will be `3`; 
  - when `FALSE`, `y` will be `NULL`; 
  - when `NA` the `if` statement will throw an error.
3. 
  - You can rewrite any `for` loop to use `while` instead, and you can rewrite any `while` loop to use `repeat`, but the converses are not true. That means `while` is more flexible than `for`, and `repeat` is more flexible than `while`.
```{r, eval = FALSE}
if("2") 3
if(2) 3

for(i in 1:3){
  print(i)
}

i <- 1
while(i < 4){
  print(i)
  i <- i + 1
}

i <- 1
repeat{
  print(i)
  i <- i + 1 
  if(i > 3) break
  
}
```
  




