---
title: "Advanced R for Econometricians"
subtitle: "Databases using R"
author: 
  - "Martin Arnold"
  - "Martin Schmelzer"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/ude_fonts.css", "../assets/ude.css", "../assets/title_slides.css", "../assets/custom.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/navigate_home.js", "../assets/remark-zoom.js", "../xaringan_files/macros.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
class: title-slide title-ggplot2 center middle
```{r setup, echo = F, eval = T, warning = F, message = F}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE, 
                      fig.align = 'center', cache = F,  
                      dpi = 250, fig.width=10, out.width='72%')

source(here::here("SoSe_2026/utils.R"))
fontawesome::fa_html_dependency()
options(htmltools.dir.version = FALSE)

library(magrittr)
library(ggplot2)
library(plotly)
library(gapminder)
```

```{r moon_reader, include=FALSE, eval = FALSE}
# setup for moon reader
# copy it to console
options(servr.interval = 0.5)
xaringan::inf_mr()
```


```{r, include = FALSE, cache = FALSE}
library(knitr)
knitr::opts_chunk$set(eval = FALSE)

default_source_hook <- knit_hooks$get("source")
knit_hooks$set(
  source = function(x, options) {
    if(is.null(options$table))
      default_source_hook(x, options)
    else {
      eval(parse(text = x)) %>%
        kable("html") %>%
        kable_styling("hover", full_width = F)
    }
  }
)
```

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`
### `r rmarkdown::metadata$author`

---
## Why Databases for Econometricians?

- **Memory Limitations**: RAM capacity is often smaller than modern datasets.
- **Persistence**: Data is stored in a structured way on disk rather than volatile memory.
- **Efficiency**: Only required aggregations or subsets are loaded into R.
- **Workflow**: R acts as an interface, while the database performs the heavy lifting (Lazy Evaluation).

---
## (Relational) Databases

.font90[
- A database is a place where data is stored and can be manipulated.
- A relational database aims to reduce redundancies by spreading the data over multiple tables and
  relate them to each other. 
]

--

.font90[.blockquote.exercise[

#### `r ICONS$desktop` Example: Redundent Data

```{r, echo = FALSE, eval = TRUE, results = 'asis', message = FALSE}
library(dplyr)
set.seed(123)
cust_names<- c("17" = "Peter", "34" = "Kelly", "108" = "Aleta")
cust_address<- c("17" = "Carnaby  St.", "34" = "Downing Rd.", "108" = "Berwick St.")
sells <- tibble(store_id = rep(1:2, each = 4), 
                    store_location   = rep(c("Baker St.", "Abbey Road"), each = 4),
                    store_mngr   = rep(c("John", "George"), each = 4),
                    cust_id      = sample(c(34, 17, 108, 17, 108, 17, 34, 17), 8, replace = TRUE),
                    cust_address = cust_address[as.character(cust_id)],
                    cust_name    = cust_names[as.character(cust_id)],
                    date = sample(seq(as.Date("2019/1/1"), as.Date("2019/3/1"), 
                               by = "day"), 8),
                    total = sample(10:20, 8, replace = TRUE)
                )

knitr::kable(sells, format = "html")
```

<br>
]]
---
#### Example of Non-Redundent Data 

- The same information would be stored in a relational database like this: 

--


```{r, eval = TRUE, echo = FALSE}
store <- tibble(store_id =  1:2, 
                store_location = c("Baker St.", "Abbey Road"),
                store_mngr   = c("John", "George")
                )

customer <- tibble(cust_id = names(cust_names),
                   cust_names = cust_names,
                   cust_address = cust_address)

sells_rel <- sells %>% select(-store_location, -store_mngr, -cust_address, -cust_name) 
```

.pull-left[
```{r, echo = FALSE, eval = TRUE}
knitr::kable(sells_rel, format = "html", caption = "Sell Records")
```
]
.pull-right[
```{r, echo = FALSE, eval = TRUE}
knitr::kable(store, format = "html", caption = "Store")
```
<br>
```{r, echo = FALSE, eval = TRUE}
knitr::kable(customer, format = "html", caption = "Customer")
```
]

<br>

--
- The relationship between the tables is defined by the `id` columns (called keys).



---
## DBI Interface and Backends


- `DBI` is the standard for communication between R and databases. 
- Provides a consistent syntax across different systems (SQL dialects).


--

**Important Backends**:

- RSQLite: File-based, ideal for local projects
- duckdb: Columnar, extremely fast for analytical queries
- RMariaDB / RPostgres: For server environments

```{r, echo = T, eval = F}
install.packages(c("DBI", "dbplyr", "RSQLite"))
```


---
## Connect to a Database


- Usually, to connect to a database authentication information such as 
  username and password are required (the database admin should provide those). 
- You also need to specify a driver that makes communication with the database possible.


--

.font80[.blockquote.exercise[

#### `r ICONS$desktop` Setup a connection 

```{r, eval = FALSE}
library(dbplyr)
library(DBI)
con <- DBI::dbConnect(
          drv = RMariaDB::MariaDB(), # MariaDB driver works for MySQL as well
          username = "admin",
          dbname   = "my_db", 
          host     = "sql11.freesqldatabase.com",
          password = "password",
          port     = "3306"
          )
```
]]





---
## Let `pool` manage your db connections

- `DBI::dbConnect()` can drop connections during unstable network conditions.
- `pool` manages connections automatically (Connection Pooling).


.blockquote.exercise[
```{r, eval = F}
install.packages("pool")
library(pool)
# Create a pool instead of a static connection
con <- pool::dbPool(
  drv = RSQLite::SQLite(),
  dbdir = "local_db.sqlite"
)

# ... Operations ...

# Close pool safely
pool::poolClose(con)
```
]






---

.font90[
## Set up a SQLite Database

- SQLite is a light weight Database Management System 
- An SQLite database is contained in a single file. There is no need for a server. 
- Everything you need is to install the `RSQLite` package. 
]

--

.font90[
#### Set up a SQLite Database
- A new SQLite database is automatically created if you call `DBI::dbConnect()` (or `dbPool()`)
with argument `dbname = "path_to_database"` where `"path_to_database"` is a path to 
a non-existing file. 
- In case `"path_to_database"` is an existing SQLite database it 
connects to that database. 
]

--

.font90.blockquote.exercise[
#### `r ICONS$desktop` Example 

```{r, echo = TRUE, eval = FALSE}
library(dplyr)
pool <- pool::dbPool(
  drv = RSQLite::SQLite(),
  dbname = "local_db.sqlite"
)
pool::poolClose(pool)
```
]

???

SQLite -> SQLite is a C-language library that implements a small, fast, self-contained, high-reliability, full-featured, SQL database engine.



---
## Write Data to the Database
.font80[
- When the connection is established we can use the object `pool` to communicate with the database.
- We add the data frame `mtcars` as a new table to the database.
]
--

.font90.blockquote.exercise[
#### `r ICONS$desktop` Example 
```{r, eval = FALSE}
pool::dbWriteTable(pool, name = "tbl_mtcars", value = mtcars)
# OR
dplyr::copy_to(pool, name = "tbl_mtcars", df = mtcars)
```
]

--
.font80[
- To add data to an existing table we use the additional argument `append = TRUE` (or `dbAppendTable`).
]

--

.font90.blockquote.exercise[
#### `r ICONS$desktop` Example 

```{r, eval = FALSE}
nrw_row <- data.frame(mpg = 50, cyl = 7, disp = 200, hp = 323, drat = 4.2, wt = 3.1, 
                      qsec = 19.2, vs = 0, am = 1, gear = 5, carb = 3)
dbWriteTable(pool, name = "tbl_mtcars", value = nrw_row, append = T)
pool::poolClose(con)
```
]

--
.font80[
- Always make sure, your pool/connection is being closed using `closePool` (or, if you used `dbConnect` directly, `dbDisconnect`)
]

???
Closing connections: pool manages connections for you

---
## Access Data

- The function `dbListTables()` tells us which tables are in the database.

.blockquote.exercise[
```{r, eval = FALSE}
dbListTables(pool)
```
]

--

- With `dplyr::tbl()` we can create an object which is a reference to a table in the database. Most `dplyr` functions work on this object as it was a (regular) data frame (or `tibble`).

--

.blockquote.exercise[
#### `r ICONS$desktop` Example 
```{r, eval = FALSE}
tbl_mtcars <- tbl(pool, "tbl_mtcars")
class(tbl_mtcars)
```
]
---
## Use Standard `dplyr`  Functions on the Database

.font90.blockquote.exercise[

#### `r ICONS$desktop` Example 
```{r, message=FALSE, warning=FALSE, eval = FALSE}
avg_mpg <- tbl_mtcars %>%
  group_by(cyl) %>% 
  select_if(is.numeric) %>% 
  summarise_all(mean)
```

]

--

.font90[
- Even if it seems as if `avg_mpg` is a data frame it still is only a set of instructions to be performed by the database. 
- `show_query()` tells us which instructions are sent to the database. 
]

--

.font90.blockquote.exercise[

#### `r ICONS$desktop` Example 
```{r, warning=FALSE, eval = FALSE}
show_query(avg_mpg)
```
]

???

- select_if superseded
- but dbplyr **do not** support predicate functions, i.e. you can not use where(is.numeric) to select all numeric variables.

---
## Use Standard `dplyr` Functions on the Database

.font90[
When working with databases, `dplyr` tries to be as lazy as possible:

- It never pulls data into R unless you explicitly ask for it.
- It delays doing any work until the last possible moment: it collects everything you want to do and then sends it to the database in one step.
- `collect()` actually pulls the data into R.  
]

--

.font90.blockquote.exercise[
#### `r ICONS$desktop` Example 

```{r, eval = FALSE}
avg_mpg <- avg_mpg %>% collect(n = Inf)
print(avg_mpg)
```
]

---
## Raw SQL

- If dbplyr reaches its limits, you can execute SQL directly.

.blockquote.exercise[
```{r, eval = F}
# Passing SQL query as a string
query <- "SELECT cyl, AVG(mpg) FROM tbl_mtcars GROUP BY cyl"
result <- pool::dbGetQuery(pool, query)

# Returns a standard R data.frame
head(result)
```
]


---
## Disconnect

- If the connection to the database is no longer needed you should disconnect to free system resources. 

--

.blockquote.exercise[
```{r, eval = FALSE}
DBI::dbDisconnect(con)
rm(con)

# OR
pool::poolClose(pool)
rm(pool) # tidy up
```
]




---
class: exercise_slide
## Exercises

1. Set up a new SQLite database and write the `gapminder::gapminder` dataset to it.

2. Add a new row of (fictive) data.

3. Add the `gapminder::country_codes` dataset to it.

4. Use `dplyr::tbl` and select all unique countries in the gapminder table.

5. Create a query to select all records where the year is 2007 and lifeExp is greater than 80. Retrieve the results.

6. Calculate the average gdpPercap per continent for the year 2002.

7. Join `gapminder` and `country_codes` by `country`. Filter for 2007 and select `country`, `year`, and `iso_alpha`.





