---
title: "Rcpp --- Solutions to Exercises"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# `Rcpp` &mdash; Exercises Part 1

- Find out why the following code gives a compile error:
    ```{Rcpp, eval=F}
    #include<Rcpp.h>
    // [[Rcpp::plugins("cpp11")]]
    
    NumericVector x{1, 2, 3, 4, 5};
    IntegerVector id{1};
    
    double y = x[id]; // produces compile error
    ```

    **Solution:**
    
    `x[id]` returns a subview class of `Rcpp::Vector` which is *not* a double. It is important that coercion to another type is not as easily done as in R!

- Benchmark the functions below against each other for `x<-rnorm(1e2)`, `x<-rnorm(1e4)` and `x<-rnorm(1e6)`. Comment on the results.
    ```{r, echo = F}
    cppFunction('
        NumericVector test_clone_return(NumericVector A) {
      NumericVector B = clone(A);
      B[1] = 0.5;
      return B;
    }
    ')
    cppFunction('
    NumericVector test_reference_return(NumericVector A) {
      A[1] = 0.5;
      return A;
    }
    ')
    ```
    
    ```{Rcpp, eval=F}
    NumericVector test_clone_return(NumericVector A) {
      NumericVector B = clone(A);
      B[1] = 0.5;
      return B;
    }
    ```
    ```{Rcpp, eval=F}
    NumericVector test_reference_return(NumericVector A) {
      A[1] = 0.5;
      return A;
    }
    ```

  **Solution:**

    ```{r}
    bench::press(
      A = c(1e2, 1e4, 1e6),
      {
        A <- rnorm(A)  
        bench::mark(
          test_clone_return(A),
          test_reference_return(A), 
          check = F,
          relative = T
        )
      }
    )
    ```
    
Obviously, cloning (that is copying on function call) is a bad idea!

# Slide 52: `Rcpp` -- STL Algorithms

The second appearence of `x.begin()` refers to the biginning of the output range, i.e. the first element of `x`.

    